---
title: "Two stage DiD"
---

$E(\varepsilon_{igt}|D_{gt}, g, t) = 0$

Static TWFE Model (Equation 1) $y_{igt} = \mu_g + \eta_t + \tau D_{gt} + \varepsilon_{igt}$

Dynamic TWFE Model $y_{igt} = \mu_g + \eta_t + \tau_{gt}D_{gt} + \varepsilon_{igt}$

Overall Average Treatment Effect $\tau \equiv \sum_{g,t} \frac{1}{N_{gt}} \tau_{gt}$

Static TWFE Equivalence (FWL Application): $y_{igt} - \hat{\mu}_g - \hat{\eta}_t = \tau \tilde{D}_{gt} + \tilde{\varepsilon}_{gt}$

-   starting with $y_{igt} = \mu_g + \eta_t + \tau_{gt}D_{gt} + \varepsilon_{igt}$ regress $y_{igt}$ on fixed effects $[μ_g, η_t]$

    -   the fitted values are $\hat{y}_{igt} = \hat{\mu}_g + \hat{\eta}_t$ and so the residuals are $\tilde{y}_{igt} = y_{igt} - \hat{\mu}_g -\hat{\eta}_t$

-   Regress the treatment indicator $D_{gt}$ on the same fixed effects $[μ_g, η_t]$

    -   the fitted values are $\hat{D}_{gt}$ and so the residuals are $\tilde{D}_{gt} = D_{gt} -\hat{D}_{gt}$

-   regress $\tilde{y}_{igt} = \tau\tilde{D}_{gt} + \tilde{\varepsilon}_{gt}$

The key insight is that $\tilde{D}_{gt}$ (the residualized treatment indicator) varies across different group-time combinations in complex ways. When there's treatment effect heterogeneity ($\tau_{gt}$ varies), the FWL theorem shows that what we're actually estimating is:

$τ̂ = Σ_{g,t} w_{gt} τ_{gt}$

FWL Theorem Expression $\hat{\tau}_{gt} = \tau \tilde{D}_{gt} + \tilde{\varepsilon}_{gt}$

# Why Residualized Treatment Indicators Create Problematic Weights

## The Residualization Process

When we apply FWL, we regress the treatment indicator $D_{gt}$ on group and time fixed effects:

$$D_{gt} = \alpha_g + \delta_t + \text{residual}$$

The residual is: $\tilde{D}_{gt} = D_{gt} - \hat{\alpha}_g - \hat{\delta}_t$

## Key Insight: Treatment Patterns Matter

Let's see why $\tilde{D}_{gt}$ varies across group-time combinations with a concrete example.

### Example: Staggered Treatment Adoption

Consider 3 groups adopting treatment in different years: - **Group A**: Treated starting year 3 - **Group B**: Treated starting year 5 - **Never-treated group**: Never treated

**Treatment Matrix (**$D_{gt}$):

$$         
Year:    1  2  3  4  5  6
Group A: 0  0  1  1  1  1
Group B: 0  0  0  0  1  1  
Never:   0  0  0  0  0  0
$$

## Step 1: Estimate Fixed Effects from Treatment Pattern

### Group Fixed Effects ($\hat{\alpha}_g$)

-   $\hat{\alpha}_A = \frac{4}{6} = 0.67$ (Group A treated 4 out of 6 periods)
-   $\hat{\alpha}_B = \frac{2}{6} = 0.33$ (Group B treated 2 out of 6 periods)\
-   $\hat{\alpha}_{Never} = 0$ (Never treated)

### Time Fixed Effects ($\hat{\delta}_t$)

-   $\hat{\delta}_1 = \hat{\delta}_2 = 0$ (No one treated in years 1-2)
-   $\hat{\delta}_3 = \frac{1}{3} = 0.33$ (1 out of 3 groups treated in year 3)
-   $\hat{\delta}_4 = \frac{1}{3} = 0.33$ (1 out of 3 groups treated in year 4)
-   $\hat{\delta}_5 = \frac{2}{3} = 0.67$ (2 out of 3 groups treated in year 5)
-   $\hat{\delta}_6 = \frac{2}{3} = 0.67$ (2 out of 3 groups treated in year 6)

## Step 1(a): Estimate Fixed Effects from Treatment Pattern

When we regress $D_{gt} = \alpha_g + \delta_t + \text{residual}$, we need **identification restrictions**.

Standard approach: Set one group FE and one time FE to zero. - Set $\alpha_{Never} = 0$ (never-treated group as base) - Set $\delta_1 = 0$ (first period as base)

The remaining fixed effects solve: $\hat{\alpha}_g + \hat{\delta}_t = \text{average of } D_{gt} \text{ for group } g \text{ in period } t$

### From our treatment matrix, the cell means are:

```         
       Year 1  Year 2  Year 3  Year 4  Year 5  Year 6
Group A:  0      0      1       1       1       1
Group B:  0      0      0       0       1       1  
Never:    0      0      0       0       0       0
```

### Solving the System:

With $\alpha_{Never} = 0$ and $\delta_1 = 0$:

-   $\hat{\alpha}_A + 0 = 0 \Rightarrow \hat{\alpha}_A = 0$
-   $0 + \hat{\delta}_2 = 0 \Rightarrow \hat{\delta}_2 = 0$\
-   $\hat{\alpha}_A + \hat{\delta}_3 = 1 \Rightarrow 0 + \hat{\delta}_3 = 1 \Rightarrow \hat{\delta}_3 = 1$
-   $\hat{\alpha}_A + \hat{\delta}_4 = 1 \Rightarrow 0 + \hat{\delta}_4 = 1 \Rightarrow \hat{\delta}_4 = 1$
-   $\hat{\alpha}_A + \hat{\delta}_5 = 1$ and $\hat{\alpha}_B + \hat{\delta}_5 = 1$

This system is **over-identified** - we can't satisfy all constraints simultaneously!

## The Real Issue: Perfect Collinearity

The problem is that **treatment status is perfectly predictable** from group and time fixed effects in many staggered designs. This creates perfect or near-perfect collinearity.

### What Actually Happens in Practice:

1.  **Regularization**: Statistical software uses generalized inverses or drops redundant parameters
2.  **Approximate relationships**: With heterogeneous treatment effects, there's variation around the fixed effect predictions
3.  **Sample variation**: Real data has idiosyncratic variation that breaks perfect collinearity

### More Realistic Example:

The residuals $\tilde{D}_{gt}$ are typically **small but non-zero** due to: - Sampling variation in treatment effects - Slight deviations from perfect treatment timing - Heterogeneity within group-time cells

The **key insight remains**: Even small residuals create problematic weights because: - Some $\tilde{D}_{gt} > 0$ for untreated units\
- Some $\tilde{D}_{gt} < 0$ for treated units - Weights become inversely related to actual treatment probabilities

## Step 2: Calculate Residualized Treatment $\tilde{D}_{gt}$

For each group-time combination: $$\tilde{D}_{gt} = D_{gt} - \hat{\alpha}_g - \hat{\delta}_t$$

### Group A Results:

```         
Year:    1     2     3     4     5     6
D_gt:    0     0     1     1     1     1
α̂_A:    0.67  0.67  0.67  0.67  0.67  0.67
δ̂_t:    0     0     0.33  0.33  0.67  0.67
D̃_gt:  -0.67 -0.67  0     0     -0.33 -0.33
```

### Group B Results:

```         
Year:    1     2     3     4     5     6
D_gt:    0     0     0     0     1     1
α̂_B:    0.33  0.33  0.33  0.33  0.33  0.33
δ̂_t:    0     0     0.33  0.33  0.67  0.67
D̃_gt:  -0.33 -0.33 -0.67 -0.67  0     0
```

### Never-Treated Group Results:

```         
Year:    1     2     3     4     5     6
D_gt:    0     0     0     0     0     0
α̂_N:    0     0     0     0     0     0
δ̂_t:    0     0     0.33  0.33  0.67  0.67
D̃_gt:   0     0    -0.33 -0.33 -0.67 -0.67
```

## Key Observations

### 1. Negative Residuals for Treated Observations

Notice that **many treated observations** (where $D_{gt} = 1$) have **negative** $\tilde{D}_{gt}$ values: - Group A in years 5-6: $\tilde{D}_{gt} = -0.33$ - This happens because these groups are "over-treated" relative to the average

### 2. Positive Residuals for Some Untreated Observations

Some **untreated observations** have **positive** $\tilde{D}_{gt}$ values: - This creates the wrong sign relationship

### 3. Magnitude Varies by Treatment Timing

-   **Early adopters** (Group A) have different residual patterns than **late adopters** (Group B)
-   **Never-treated** units have consistently negative residuals in later periods

## The Weighting Problem

When we run the second-stage regression: $$\tilde{y}_{igt} = \tau \tilde{D}_{gt} + \tilde{\varepsilon}_{gt}$$

The weight on each $\tau_{gt}$ in the final estimate $\hat{\tau}$ is proportional to: $$w_{gt} \propto \text{Var}(\tilde{D}_{gt}) \times N_{gt}$$

### Problems:

1.  **Negative weights**: Groups with negative $\tilde{D}_{gt}$ get negative weights
2.  **Unequal weights**: Groups treated at different times get very different weights
3.  **Wrong magnitudes**: The weights don't reflect the actual sample sizes or treatment intensities

## Why Gardner's Method Fixes This

Gardner's two-stage approach: 1. **Estimates fixed effects using only untreated observations** ($D_{gt} = 0$) 2. **Never residualizes the treatment indicator** 3. **Uses original** $D_{gt}$ in second stage, giving equal weight ($\frac{1}{N_{gt}}$) to each group-time combination

This ensures: $$\hat{\tau} = \sum_{g,t} \frac{1}{N_{gt}} \tau_{gt}$$

Rather than the problematic: $$\hat{\tau} = \sum_{g,t} w_{gt} \tau_{gt} \quad \text{where some } w_{gt} < 0$$

2x2 (two groups, two periods)

```{r}
#| eval: false
# Simple 2x2 Example: Traditional TWFE vs Gardner's Method
# 2 groups, 2 periods, one group treated only in period 2

library(fixest)
library(dplyr)

# =============================================================================
# CREATE SIMPLE 2x2 DATA
# =============================================================================

# Create the classic 2x2 difference-in-differences setup
data <- data.frame(
  group = c("control", "control", "treatment", "treatment"),
  period = c(1, 2, 1, 2),
  treated = c(0, 0, 0, 1),
  # Create outcomes with known treatment effect
  outcome = c(10, 11, 12, 16)  # Treatment effect = 3
)

# Add group and period as factors for fixed effects
data$group_id <- ifelse(data$group == "control", 1, 2)
data$period_id <- data$period

print("Data:")
print(data)

# =============================================================================
# CALCULATE TRUE TREATMENT EFFECT MANUALLY
# =============================================================================

# Classic DiD calculation: (Y_treat_post - Y_treat_pre) - (Y_control_post - Y_control_pre)
true_effect <- (16 - 12) - (11 - 10)
cat("\nTRUE Treatment Effect (manual DiD):", true_effect, "\n\n")

# =============================================================================
# METHOD 1: TRADITIONAL TWFE (WRONG!)
# =============================================================================

cat("=== TRADITIONAL TWFE ESTIMATION ===\n")

# This applies FWL and gives wrong answer due to perfect collinearity
traditional_twfe <- feols(outcome ~ treated | group_id + period_id, data = data)

cat("Traditional TWFE Estimate:", coef(traditional_twfe)["treated"], "\n")
cat("Standard Error:", se(traditional_twfe)[1], "\n\n")

# Let's see what's happening under the hood with FWL
cat("--- Understanding the FWL Problem ---\n")

# Step 1: Regress treatment on fixed effects
treatment_on_fe <- feols(treated ~ 1 | group_id + period_id, data = data)
data$treated_resid <- resid(treatment_on_fe)

cat("Residualized treatment (D_tilde):\n")
print(data[, c("group", "period", "treated", "treated_resid")])

# Step 2: Regress outcome on fixed effects  
outcome_on_fe <- feols(outcome ~ 1 | group_id + period_id, data = data)
data$outcome_resid <- resid(outcome_on_fe)

cat("\nResidualized outcomes (y_tilde):\n")
print(data[, c("group", "period", "outcome", "outcome_resid")])

# Step 3: Regress residualized outcome on residualized treatment
fwl_manual <- lm(outcome_resid ~ treated_resid - 1, data = data)
cat("\nManual FWL estimate:", coef(fwl_manual)[1], "\n")
cat("(Should match traditional TWFE above)\n\n")

# Show the problem: which observations drive identification?
cat("Key insight: Only observations with non-zero treated_resid matter!\n")
contributing_obs <- data[abs(data$treated_resid) > 1e-10, ]
print(contributing_obs[, c("group", "period", "treated", "treated_resid", "outcome_resid")])

resid(outcome_on_fe) |> matrix(nrow=2)
outcome_on_fe$fitted.values |> matrix(nrow=2)
(outcome_on_fe$fitted.values -outcome_on_fe$fitted.values[1]) |> matrix(nrow=2)

resid(treatment_on_fe) |> matrix(nrow=2)
treatment_on_fe$fitted.values |> matrix(nrow=2)
(treatment_on_fe$fitted.values -treatment_on_fe$fitted.values[1]) |> matrix(nrow=2)


# =============================================================================
# METHOD 2: GARDNER'S TWO-STAGE METHOD (CORRECT!)
# =============================================================================

cat("\n=== GARDNER'S TWO-STAGE METHOD ===\n")

# Stage 1: Estimate fixed effects using ONLY UNTREATED observations
untreated_data <- data[data$treated == 0, ]
cat("Stage 1 data (untreated only):\n")
print(untreated_data[, c("group", "period", "outcome")])

stage1 <- feols(outcome ~ 1 | group_id + period_id, data = untreated_data)

# Get predicted values for all observations
data$outcome_pred_stage1 <- predict(stage1, newdata = data)
data$outcome_resid_gardner <- data$outcome - data$outcome_pred_stage1



cat("\nStage 1 predictions and residuals:\n")
print(data[, c("group", "period", "outcome", "outcome_pred_stage1", "outcome_resid_gardner")])

# Stage 2: Regress residualized outcome on ORIGINAL treatment indicator
stage2 <- feols(outcome_resid_gardner ~ treated, data = data)

cat("\nGardner Two-Stage Estimate:", coef(stage2)[1], "\n")
cat("Standard Error:", se(stage2)[1], "\n\n")

# =============================================================================
# MANUAL CALCULATION TO VERIFY GARDNER'S METHOD
# =============================================================================

cat("=== MANUAL VERIFICATION ===\n")

# Stage 1 manual calculation
control_period1 <- 10
control_period2 <- 11
treatment_period1 <- 12

# Estimate fixed effects from untreated obs
group_effect_control <- 0  # normalization
group_effect_treatment <- treatment_period1 - control_period1  # 12 - 10 = 2
time_effect_period1 <- 0   # normalization  
time_effect_period2 <- control_period2 - control_period1      # 11 - 10 = 1

cat("Estimated fixed effects (from untreated obs only):\n")
cat("Group effect (treatment):", group_effect_treatment, "\n")
cat("Time effect (period 2):", time_effect_period2, "\n")

# Predict counterfactual for treated unit in period 2
counterfactual_treated_period2 <- group_effect_treatment + time_effect_period2 + control_period1
# = 2 + 1 + 10 = 13

actual_treated_period2 <- 16
manual_treatment_effect <- actual_treated_period2 - counterfactual_treated_period2

cat("Counterfactual for treated unit in period 2:", counterfactual_treated_period2, "\n")
cat("Actual outcome for treated unit in period 2:", actual_treated_period2, "\n")
cat("Manual Gardner estimate:", manual_treatment_effect, "\n\n")

# =============================================================================
# SUMMARY COMPARISON
# =============================================================================

cat("=== SUMMARY ===\n")
cat("True treatment effect:          ", true_effect, "\n")
cat("Traditional TWFE estimate:      ", round(coef(traditional_twfe)[1], 3), "\n") 
cat("Gardner two-stage estimate:     ", round(coef(stage2)[1], 3), "\n")
cat("Manual Gardner calculation:     ", manual_treatment_effect, "\n\n")

if (abs(coef(traditional_twfe)[1] - true_effect) > 0.001) {
  cat("❌ Traditional TWFE is WRONG!\n")
} else {
  cat("✅ Traditional TWFE is correct\n")
}

if (abs(coef(stage2)[1] - true_effect) < 0.001) {
  cat("✅ Gardner's method is CORRECT!\n")
} else {
  cat("❌ Gardner's method is wrong\n")
}

# =============================================================================
# SHOW WHY TRADITIONAL TWFE FAILS
# =============================================================================

cat("\n=== WHY TRADITIONAL TWFE FAILS ===\n")
cat("The problem: Perfect collinearity in 2x2 design\n")
cat("Treatment is perfectly predicted by group + time fixed effects\n")
cat("This makes the residualized treatment indicator problematic\n\n")

cat("Residualized treatment values:\n")
for(i in 1:nrow(data)) {
  cat(sprintf("Group %s, Period %d: D_tilde = %.3f\n", 
              data$group[i], data$period[i], data$treated_resid[i]))
}

cat("\nNotice: The actually treated observation (treatment, period 2) has D_tilde ≈ 0\n")
cat("So it contributes almost nothing to the TWFE estimate!\n")
cat("Identification comes from the wrong observations.\n")
```

3x3 (3 groups, 3 periods, homogeneous)

```{r}
#| eval: false
# Staggered Treatment with HOMOGENEOUS Effects: TWFE vs Gardner
# Testing whether TWFE works when treatment effects are constant across groups/time

library(fixest)
library(dplyr)

# =============================================================================
# CREATE STAGGERED DATA WITH HOMOGENEOUS (CONSTANT) TREATMENT EFFECTS
# =============================================================================

data <- data.frame(
  group = rep(c("A", "B", "C"), each = 3),
  period = rep(1:3, 3),
  group_id = rep(c(1, 2, 3), each = 3),
  period_id = rep(1:3, 3)
) %>%
  dplyr::mutate(
    # Treatment indicator (staggered timing)
    treated = case_when(
      group == "A" & period >= 2 ~ 1,
      group == "B" & period >= 3 ~ 1,
      TRUE ~ 0
    )
  )

# Create outcomes with HOMOGENEOUS (constant) treatment effects
data <- data %>%
  dplyr::mutate(
    # Base outcome (group and time fixed effects)
    base_outcome = case_when(
      group == "A" ~ 10,  # Group A baseline
      group == "B" ~ 12,  # Group B baseline  
      group == "C" ~ 8    # Group C baseline
    ) + (period - 1) * 1.5,  # Common time trend
    
    # HOMOGENEOUS treatment effect: same for all treated units!
    treatment_effect = ifelse(treated == 1, 3.0, 0),  # Constant effect of 3.0
    
    # Final outcome
    outcome = base_outcome + treatment_effect
  )

print("Data with HOMOGENEOUS treatment effects:")
print(data[, c("group", "period", "treated", "treatment_effect", "outcome")])

# =============================================================================
# TRUE TREATMENT EFFECT (should be 3.0 for all)
# =============================================================================

true_effects <- data %>%
  filter(treated == 1) %>%
  select(group, period, treatment_effect)

true_att <- mean(true_effects$treatment_effect)
cat("\nTRUE Average Treatment Effect:", true_att, "\n")
cat("Individual treatment effects (all the same):\n")
print(true_effects)

# =============================================================================
# METHOD 1: TRADITIONAL TWFE
# =============================================================================

cat("\n=== TRADITIONAL TWFE ESTIMATION ===\n")

traditional_twfe <- feols(outcome ~ treated | group_id + period_id, data = data)

cat("Traditional TWFE Estimate:", round(coef(traditional_twfe)[1], 3), "\n")
cat("Standard Error:", round(se(traditional_twfe)[1], 3), "\n")

# Analyze the weighting
treatment_on_fe <- feols(treated ~ 1 | group_id + period_id, data = data)
data$treated_resid <- resid(treatment_on_fe)

outcome_on_fe <- feols(outcome ~ 1 | group_id + period_id, data = data)
data$outcome_resid <- resid(outcome_on_fe)

lm(outcome_resid ~ treated_resid, data = data_h3)

cat("\nResidualized treatment and outcomes:\n")
print(data[, c("group", "period", "treated", "treated_resid", "outcome_resid")])

# Calculate implicit weights
nonzero_resid <- data[abs(data$treated_resid) > 1e-10, ]
if(nrow(nonzero_resid) > 0) {
  weights <- nonzero_resid$treated_resid^2 / sum(nonzero_resid$treated_resid^2)
  
  cat("\nImplicit weights in TWFE estimation:\n")
  weight_table <- data.frame(
    group = nonzero_resid$group,
    period = nonzero_resid$period,
    treated = nonzero_resid$treated,
    treated_resid = round(nonzero_resid$treated_resid, 3),
    weight = round(weights, 3),
    true_effect = nonzero_resid$treatment_effect
  )
  print(weight_table)
  
  # Since all treatment effects are the same, weighted average = unweighted average
  weighted_estimate <- sum(weights * weight_table$true_effect)
  cat("\nWeighted average using TWFE weights:", round(weighted_estimate, 3), "\n")
}

# =============================================================================
# METHOD 2: GARDNER'S TWO-STAGE METHOD
# =============================================================================

cat("\n=== GARDNER'S TWO-STAGE METHOD ===\n")

# Stage 1: Estimate fixed effects using ONLY UNTREATED observations
untreated_data <- data[data$treated == 0, ]
stage1 <- feols(outcome ~ 1 | group_id + period_id, data = untreated_data)

# Stage 2: Regress residualized outcome on treatment
data$outcome_resid_gardner <- data$outcome - predict(stage1, newdata = data)
stage2 <- feols(outcome_resid_gardner ~ treated, data = data)

cat("Gardner Two-Stage Estimate:", round(coef(stage2)[1], 3), "\n")
cat("Standard Error:", round(se(stage2)[2], 3), "\n")

# =============================================================================
# COMPARE WITH HETEROGENEOUS EFFECTS CASE
# =============================================================================

cat("\n=== COMPARISON: HOMOGENEOUS vs HETEROGENEOUS EFFECTS ===\n")

# Now create the same data but with heterogeneous effects
data_hetero <- data %>%
  dplyr::mutate(
    # HETEROGENEOUS treatment effects (like previous example)
    treatment_effect_hetero = dplyr::case_when(
      group == "A" & period == 2 ~ 3.0,  # Same timing, different effects
      group == "A" & period == 3 ~ 4.0,  
      group == "B" & period == 3 ~ 2.0,  
      TRUE ~ 0
    ),
    outcome_hetero = base_outcome + treatment_effect_hetero
  )

# Traditional TWFE with heterogeneous effects
twfe_hetero <- feols(outcome_hetero ~ treated | group_id + period_id, data = data_hetero)

# Gardner with heterogeneous effects
data_hetero$outcome_resid_hetero <- data_hetero$outcome_hetero - predict(stage1, newdata = data_hetero)
gardner_hetero <- feols(outcome_resid_hetero ~ treated, data = data_hetero)

true_att_hetero <- mean(data_hetero[data_hetero$treated == 1, "treatment_effect_hetero"])

cat("HOMOGENEOUS Effects:\n")
cat("  True effect:          ", round(true_att, 3), "\n")
cat("  Traditional TWFE:     ", round(coef(traditional_twfe)[1], 3), "\n")
cat("  Gardner method:       ", round(coef(stage2)[3], 3), "\n\n")

cat("HETEROGENEOUS Effects:\n")
cat("  True average effect:  ", round(true_att_hetero, 3), "\n")
cat("  Traditional TWFE:     ", round(coef(twfe_hetero)[1], 3), "\n")
cat("  Gardner method:       ", round(coef(gardner_hetero)[2], 3), "\n\n")

# =============================================================================
# KEY INSIGHT
# =============================================================================

cat("=== KEY INSIGHT ===\n")

homo_twfe_error <- abs(coef(traditional_twfe)[1] - true_att)
homo_gardner_error <- abs(coef(stage2)[1] - true_att)

hetero_twfe_error <- abs(coef(twfe_hetero)[1] - true_att_hetero)
hetero_gardner_error <- abs(coef(gardner_hetero)[1] - true_att_hetero)

cat("Errors with HOMOGENEOUS effects:\n")
cat("  TWFE error:           ", round(homo_twfe_error, 3), "\n")
cat("  Gardner error:        ", round(homo_gardner_error, 3), "\n\n")

cat("Errors with HETEROGENEOUS effects:\n")
cat("  TWFE error:           ", round(hetero_twfe_error, 3), "\n")
cat("  Gardner error:        ", round(hetero_gardner_error, 3), "\n\n")

if (homo_twfe_error < 0.1 && homo_gardner_error < 0.1) {
  cat("✅ With HOMOGENEOUS effects: Both methods work!\n")
} else {
  cat("❌ Problem even with homogeneous effects\n")
}

if (hetero_twfe_error > 0.1 && hetero_gardner_error < 0.1) {
  cat("❌ With HETEROGENEOUS effects: Only Gardner works!\n")
} else if (hetero_twfe_error < 0.1 && hetero_gardner_error < 0.1) {
  cat("✅ With HETEROGENEOUS effects: Both methods work\n")
}

cat("\nCONCLUSION: The TWFE weighting problem only matters when treatment\n")
cat("effects are HETEROGENEOUS across groups or time periods.\n")
```

Diagnostic

```{r}
#| eval: false
# Diagnostic: Understanding why Gardner's method gives 0 instead of 3

library(fixest)
library(dplyr)

# Simple 3x3 homogeneous case
data <- data.frame(
  group = rep(c("A", "B", "C"), each = 3),
  period = rep(1:3, 3),
  group_id = rep(c(1, 2, 3), each = 3)
) %>%
  dplyr::mutate(
    treated = dplyr::case_when(
      group == "A" & period >= 2 ~ 1,
      group == "B" & period >= 3 ~ 1,
      TRUE ~ 0
    ),
    # Simple structure: group effects + time effects + treatment effect
    outcome = dplyr::case_when(
      group == "A" ~ 10,
      group == "B" ~ 12,  
      group == "C" ~ 8
    ) + (period - 1) * 1.5 + ifelse(treated == 1, 3.0, 0)
  )

print("Data:")
print(data)

cat("\n=== STEP-BY-STEP DIAGNOSTIC ===\n")

# Traditional TWFE
twfe <- feols(outcome ~ treated | group_id + period, data = data)
cat("Traditional TWFE:", coef(twfe)[1], "\n\n")

# Gardner's method step by step
cat("Gardner's Method:\n")
cat("Step 1: Untreated observations only:\n")
untreated <- data[data$treated == 0, ]
print(untreated)

cat("\nStep 1 regression (outcome ~ 1 | group_id + period on untreated):\n")
stage1 <- feols(outcome ~ 1 | group_id + period, data = untreated)

# Show what stage1 estimates
cat("Stage 1 coefficients (group effects):\n")
print(fixef(stage1)$group_id)
cat("Stage 1 coefficients (period effects):\n") 
print(fixef(stage1)$period)

cat("\nPredictions for all observations:\n")
data$pred <- predict(stage1, newdata = data)
data$resid <- data$outcome - data$pred
print(data[, c("group", "period", "treated", "outcome", "pred", "resid")])

cat("\nStep 2: Regress residuals on treatment:\n")
stage2 <- feols(resid ~ treated, data = data)
cat("Gardner estimate:", coef(stage2)[1], "\n\n")

# Manual calculation to verify
cat("=== MANUAL VERIFICATION ===\n")
treated_obs <- data[data$treated == 1, ]
cat("Treated observations residuals:\n")
print(treated_obs[, c("group", "period", "outcome", "pred", "resid")])
cat("Average residual for treated obs:", mean(treated_obs$resid), "\n")

untreated_obs <- data[data$treated == 0, ]
cat("Average residual for untreated obs:", mean(untreated_obs$resid), "\n")

cat("\nDifference (treated - untreated residuals):", 
    mean(treated_obs$resid) - mean(untreated_obs$resid), "\n")
```

# The TWFE Decomposition

## Starting Point: FWL Second-Stage Regression

Starting with the FWL second-stage regression:

$$
\tilde{y}_{igt} = \tau \tilde{D}_{gt} + \tilde{\varepsilon}_{igt}
$$

The OLS estimator is:

$$
\hat{\tau}_{TWFE} = \frac{\sum_{i,g,t} \tilde{D}_{gt} \tilde{y}_{igt}}{\sum_{i,g,t} \tilde{D}_{gt}^2}
$$

## Decomposition by Group-Time Cells

We can rewrite this by grouping observations within each $(g,t)$ cell:

$$
\hat{\tau}_{TWFE} = \frac{\sum_{g,t} \tilde{D}_{gt} \sum_{i=1}^{N_{gt}} \tilde{y}_{igt}}{\sum_{g,t} N_{gt} \tilde{D}_{gt}^2}
$$

$$
= \frac{\sum_{g,t} N_{gt} \tilde{D}_{gt} \bar{\tilde{y}}_{gt}}{\sum_{g,t} N_{gt} \tilde{D}_{gt}^2}
$$

Where the average residualized outcome in cell $(g,t)$ is:

$$
\bar{\tilde{y}}_{gt} = \frac{1}{N_{gt}} \sum_{i=1}^{N_{gt}} \tilde{y}_{igt}
$$

## Weighted Average Form

This can be expressed as:

$$
\hat{\tau}_{TWFE} = \sum_{g,t} w_{gt} \bar{\tilde{y}}_{gt}
$$

Where the weights are:

$$
w_{gt} = \frac{N_{gt} \tilde{D}_{gt}}{\sum_{g',t'} N_{g't'} \tilde{D}_{g't'}}
$$

**Note:** These weights can be **negative** when $\tilde{D}_{gt} < 0$, and they sum to 1:

$$
\sum_{g,t} w_{gt} = 1
$$

## Connection to True Treatment Effects

Under parallel trends, $\bar{\tilde{y}}_{gt} \approx \tau_{gt}$ for treated cells and $\bar{\tilde{y}}_{gt} \approx 0$ for untreated cells. Therefore:

$$
\hat{\tau}_{TWFE} \approx \sum_{g,t: D_{gt}=1} w_{gt} \tau_{gt}
$$

## The Problem

The issue is that the weights $w_{gt}$ depend on the residualized treatment indicators $\tilde{D}_{gt}$, not on the natural sample sizes or treatment intensities. This leads to the problematic weighting identified in the modern DiD literature, where:

1.  Some weights can be **negative**
2.  Weights don't correspond to **intuitive sample-size weighting**
3.  The estimate can have the **wrong sign** relative to the true average treatment effect

## Gardner's Solution

Gardner's method avoids this by ensuring:

$$
\hat{\tau}_{Gardner} = \sum_{g,t: D_{gt}=1} \frac{1}{N_{treated}} \tau_{gt}
$$

Where each treated group-time combination gets equal weight $\frac{1}{N_{treated}}$ in the final average.

# Derivation: Residualized Outcomes Under Parallel Trends

## Starting Assumptions

### The True Data Generating Process

Under parallel trends, outcomes follow: $$
y_{igt} = \mu_g + \eta_t + \tau_{gt} D_{gt} + \varepsilon_{igt}
$$

Where:

-   $\mu_g$ = group fixed effect -

-   $\eta_t$ = time fixed effect

-   $\tau_{gt}$ = group-time specific treatment effect

-   $\varepsilon_{igt}$ = idiosyncratic error with $E[\varepsilon_{igt}] = 0$

### The FWL First Stage

In the FWL first stage, we regress $y_{igt}$ on group and time fixed effects using **all observations**: $$
y_{igt} = \hat{\mu}_g + \hat{\eta}_t + \text{residual}_{igt}
$$

The residualized outcome is: $$
\tilde{y}_{igt} = y_{igt} - \hat{\mu}_g - \hat{\eta}_t
$$

## Derivation for Different Cases

### Case 1: Untreated Cells ($D_{gt} = 0$)

For untreated observations, the true DGP becomes: $$
y_{igt} = \mu_g + \eta_t + \varepsilon_{igt}
$$

Under standard fixed effects estimation, we have: $$
\hat{\mu}_g + \hat{\eta}_t \approx \mu_g + \eta_t
$$ (asymptotically, when the fixed effects are well-identified)

Therefore: $$
\tilde{y}_{igt} = y_{igt} - \hat{\mu}_g - \hat{\eta}_t \approx (\mu_g + \eta_t + \varepsilon_{igt}) - (\mu_g + \eta_t) = \varepsilon_{igt}
$$

Taking the average within cell $(g,t)$: $$
\bar{\tilde{y}}_{gt} = \frac{1}{N_{gt}} \sum_{i=1}^{N_{gt}} \tilde{y}_{igt} \approx \frac{1}{N_{gt}} \sum_{i=1}^{N_{gt}} \varepsilon_{igt} \approx 0
$$

**Result for untreated cells:** $\bar{\tilde{y}}_{gt} \approx 0$

### Case 2: Treated Cells ($D_{gt} = 1$)

For treated observations, the true DGP is:

$$
y_{igt} = \mu_g + \eta_t + \tau_{gt} + \varepsilon_{igt}
$$

The key insight is that the fixed effects $\hat{\mu}_g$ and $\hat{\eta}_t$ are estimated from **both treated and untreated observations**. Under parallel trends, these estimates capture the **counterfactual** group and time effects that would apply to treated units in the absence of treatment.

Specifically: $$
\hat{\mu}_g + \hat{\eta}_t \approx \mu_g + \eta_t
$$

Therefore: $$
\tilde{y}_{igt} = y_{igt} - \hat{\mu}_g - \hat{\eta}_t \approx (\mu_g + \eta_t + \tau_{gt} + \varepsilon_{igt}) - (\mu_g + \eta_t) = \tau_{gt} + \varepsilon_{igt}
$$

Taking the average within cell $(g,t)$: $$
\bar{\tilde{y}}_{gt} = \frac{1}{N_{gt}} \sum_{i=1}^{N_{gt}} \tilde{y}_{igt} \approx \frac{1}{N_{gt}} \sum_{i=1}^{N_{gt}} (\tau_{gt} + \varepsilon_{igt}) = \tau_{gt} + \frac{1}{N_{gt}} \sum_{i=1}^{N_{gt}} \varepsilon_{igt} \approx \tau_{gt}
$$

**Result for treated cells:** $\bar{\tilde{y}}_{gt} \approx \tau_{gt}$

## The Parallel Trends Assumption is Crucial

The key step in both derivations is that: $$
\hat{\mu}_g + \hat{\eta}_t \approx \mu_g + \eta_t
$$

This relies on the **parallel trends assumption**: the fixed effects estimated from the data (including both treated and untreated observations) provide valid counterfactuals for what the treated units' outcomes would have been without treatment.

If parallel trends fails, then $\hat{\mu}_g + \hat{\eta}_t$ would not equal the true counterfactual $\mu_g + \eta_t$ for treated units, and the approximations would break down.

## Summary

Under parallel trends: - **Untreated cells**: $\bar{\tilde{y}}_{gt} \approx 0$ (residuals average to zero) - **Treated cells**: $\bar{\tilde{y}}_{gt} \approx \tau_{gt}$ (residuals capture the treatment effect)

This is why the TWFE estimate can be written as: $$
\hat{\tau}_{TWFE} \approx \sum_{g,t: D_{gt}=1} w_{gt} \tau_{gt}
$$

The residualized outcomes for treated cells directly estimate the group-time treatment effects, while untreated cells contribute essentially zero to the estimate.
