---
title: "TwoStageDiD_draft"
---

::: cell
```{css}
#| echo: false
pre code {
  font-size: 0.7em;
}
```
:::

## Starting Assumptions

### The True Data Generating Process

Under parallel trends, outcomes follow: $$
y_{igt} = \mu_g + \eta_t + \tau_{gt} D_{gt} + \varepsilon_{igt}
$$

Where:

-   $\mu_g$ = group fixed effect
-   $\eta_t$ = time fixed effect
-   $D_{gt}\equiv \textbf{1}_{g\le t}$
-   $\tau_{gt}$ = group-time specific treatment effect
-   $\varepsilon_{igt}$ = idiosyncratic error with $E[\varepsilon_{igt}] = 0$

## Simplest case:

$2\times 2$ DiD: two groups, two periods ($t\in\{1,2\}$); one group (A aka *control* ; $g=\infty$) never treated, the other group (B aka *treatment* ; $g=2$) treated in the second period.

Under the assumptions:

1.  **parallel trends**: that the untreated potential outcomes for both groups have parallel trends.
2.  **no anticipation**: treatment effects are zero in all periods before treatment actually begins.

then $\tau_{gt}$ represents a causal treatment effect.

In this case, one estimator is "four means and three subtractions", the reason this method is called diff-in-diff.

```{r}
#| echo: true
#| code-fold: true
#| code-summary: classic 2x2 example data
#| label: classic 2x2 example data
# Create the classic 2x2 difference-in-differences setup
data <- data.frame(
  group = c("control", "control", "treatment", "treatment"),
  period = c(1, 2, 1, 2),
  treated = c(0, 0, 0, 1),
  # Create outcomes with known treatment effect
  outcome = c(10, 11, 12, 16)  # Treatment effect = 3
)

data$group_id <- ifelse(data$group == "control", 1, 2)
data$period_id <- data$period

data |> gt::gt() |> gtExtras::gt_theme_espn()
```

with matrix representation:

```{r}
#| echo: true
#| code-fold: true
#| code-summary: classic 2x2 data matrix
#| label: classic 2x2 data matrix
data$outcome |> 
  matrix(nrow=2, dimnames = list(c("A","B"),c("1","2")))
```

The estimator is calculated by taking the means of each group, subtracting the *pre* means from the *post* means of each group, and then subtracting the treated differences from the untreated differences:

```{r}
#| echo: true
#| code-fold: true
#| code-summary: classic 2x2 four means and three subtractions
#| label: classic 2x2 four means and three subtractions
# Classic DiD calculation: 
# (Y_treat_post - Y_treat_pre) - (Y_control_post - Y_control_pre)
true_effect <- (16 - 12) - (11 - 10)
cat("\nTRUE Treatment Effect (manual DiD):", true_effect, "\n\n")
```

The fixed effects regression gives the same estimate (The `fixest` package offers a family of functions to perform estimations with multiple fixed-effects):

```{r}
#| echo: true
#| code-fold: true
#| code-summary: classic 2x2 by TWFE
#| label: classic 2x2 by TWFE

traditional_twfe <- fixest::feols(outcome ~ treated | group_id + period_id, data = data)

cat("Traditional TWFE Estimate:", coef(traditional_twfe)["treated"], "\n")
```

Finally, we use the Frisch-Waugh-Lovell (FWL) Theorem to estimate the TWFE model.

Per the FWL Theorem, the TWFE model is equivalent to the residualized regression: $y_{igt} - \hat{\mu}_g - \hat{\eta}_t = \tau \tilde{D}_{gt} + \tilde{\varepsilon}_{gt}$

-   starting with $y_{igt} = \mu_g + \eta_t + \tau_{gt}D_{gt} + \varepsilon_{igt}$ regress $y_{igt}$ on fixed effects (say $[\mu_g, \eta_t]$)
    -   the fitted values are $\hat{y}_{igt} = \hat{\mu}_g + \hat{\eta}_t$ and so the residuals are $\tilde{y}_{igt} = y_{igt} - \hat{\mu}_g -\hat{\eta}_t$
-   Regress the treatment indicator $D_{gt}$ on fixed effects (say $[\alpha_g, \delta_t]$)
    -   the fitted values are $\hat{D}_{gt}$ and so the residuals are $\tilde{D}_{gt} = D_{gt} -\hat{D}_{gt}$
-   regress $\tilde{y}_{igt} = \tau\tilde{D}_{gt} + \tilde{\varepsilon}_{gt}$

```{r}
#| echo: true
#| code-fold: true
#| code-summary: Step 1 - Regress outcome on fixed effects
#| label: Step 1 - Regress outcome on fixed effects
outcome_on_fe <- 
  fixest::feols(outcome ~ 1 | group_id + period_id, data = data)
data$outcome_resid <- resid(outcome_on_fe)
```

:::::: columns
::: {.column width="45%"}
```{r}
#| echo: true
#| code-fold: true
#| code-summary: Step 1 - outcome fitted values
#| label: Step 1 - outcome fitted values
outcome_on_fe$fitted.values |>
  matrix(nrow=2, dimnames = list(c("A","B"),c("1","2")))
```
:::

::: {.column width="10%"}
<!-- empty column to create gap -->
:::

::: {.column width="45%"}
```{r}
#| echo: true
#| code-fold: true
#| code-summary: Step 1 - outcome fitted residuals
#| label: Step 1 - outcome fitted residuals
resid(outcome_on_fe) |> 
  matrix(nrow=2, dimnames = list(c("A","B"),c("1","2")))
```
:::
::::::

The fitted values returned by `fixest::feols` do not fully identify the fixed effects, since we don't have the intercept values from the regression. This leaves us with four equations and six unknowns. So we make the following identification assumptions

$$
\mu_{\infty} = 0\;\text{and}\;\eta_1 = 0
$$

Under these assumption we can force the unit level fixed effect for the untreated units to be zero by subtracting $\hat{y}_{\infty 1}$ from the fitted values:

```{r}
#| echo: true
#| code-fold: true
#| code-summary: Step 1 - forcing mu_{infty} = 0
#| label: Step 1 - forcing mu_{infty} = 0
(outcome_on_fe$fitted.values - outcome_on_fe$fitted.values[1]) |>
  matrix(nrow=2, dimnames = list(c("A","B"),c("1","2")))
```

then, given $\eta_1 = 0$ , we have $\mu_2=2.5$. Finally, since $\mu_\infty=0$ we have $\eta_2=3.5$.

We can now compute the treatment residuals the same way.

We want to compute

$$\tilde{D}_{gt} = D_{gt} - \hat{\alpha}_g - \hat{\delta}_t$$

and then we can regress the outcome residuals on the treatment residuals, per FWL

```{r}
#| echo: true
#| code-fold: true
#| code-summary: Step 2 - Regress outcome residuals on treatment residuals
#| label: Step 2 - Regress outcome residuals on treatment residuals
#| warning: false
# regress treatment on fixed effects
treatment_on_fe <- 
  fixest::feols(treated ~ 1 | group_id + period_id, data = data)
data$treated_resid <- resid(treatment_on_fe)

# use residuals to estimate TWFE regression
lm(outcome_resid ~ treated_resid, data = data) |> broom::tidy()
```

And we see that the answer is the same as computers from the other two estimators.

Let look a little deeper at how the result was arrived at with the FWL estimator.

## Outcome Residuals in Terms of Potential Outcomes

### Setup: The Data Generating Process

Under the potential outcomes framework, each unit has two potential outcomes: - $Y_{igt}^0$ = potential outcome under no treatment - $Y_{igt}^1$ = potential outcome under treatment

The observed outcome is: $$
Y_{igt} = Y_{igt}^0 + D_{gt}(Y_{igt}^1 - Y_{igt}^0) = Y_{igt}^0 + D_{gt}\tau_{igt}
$$

Where $\tau_{igt} = Y_{igt}^1 - Y_{igt}^0$ is the individual treatment effect.

### Parallel Trends Assumption

Under parallel trends, the untreated potential outcomes follow: $$
Y_{igt}^0 = \mu_g + \eta_t + \varepsilon_{igt}
$$

Where: - $\mu_g$ = group fixed effect (time-invariant group characteristics) - $\eta_t$ = time fixed effect (group-invariant time shocks) - $\varepsilon_{igt}$ = idiosyncratic error with $E[\varepsilon_{igt}] = 0$

### The FWL First Stage: Outcome on Fixed Effects

When we regress outcomes on fixed effects using **all observations**: $$
Y_{igt} = \hat{\mu}_g + \hat{\eta}_t + \text{residual}_{igt}
$$

The outcome residuals are: $$
\tilde{Y}_{igt} = Y_{igt} - \hat{\mu}_g - \hat{\eta}_t
$$

### Decomposition by Treatment Status

#### Case 1: Untreated Observations ($D_{gt} = 0$)

For untreated units, the observed outcome is: $$
Y_{igt} = Y_{igt}^0 = \mu_g + \eta_t + \varepsilon_{igt}
$$

Under consistent fixed effects estimation: $$
\hat{\mu}_g \xrightarrow{p} \mu_g \quad \text{and} \quad \hat{\eta}_t \xrightarrow{p} \eta_t
$$

Therefore, the outcome residual is: $$
\tilde{Y}_{igt} = Y_{igt} - \hat{\mu}_g - \hat{\eta}_t \xrightarrow{p} (\mu_g + \eta_t + \varepsilon_{igt}) - (\mu_g + \eta_t) = \varepsilon_{igt}
$$

**For untreated units:** $\tilde{Y}_{igt} \approx \varepsilon_{igt}$ (idiosyncratic error)

#### Case 2: Treated Observations ($D_{gt} = 1$)

For treated units, the observed outcome is: $$
Y_{igt} = Y_{igt}^1 = Y_{igt}^0 + \tau_{igt} = \mu_g + \eta_t + \varepsilon_{igt} + \tau_{igt}
$$

The crucial insight is that $\hat{\mu}_g$ and $\hat{\eta}_t$ are estimated from **both treated and untreated observations**. Under parallel trends, these estimates converge to the **counterfactual** fixed effects: $$
\hat{\mu}_g \xrightarrow{p} \mu_g \quad \text{and} \quad \hat{\eta}_t \xrightarrow{p} \eta_t
$$

Therefore, the outcome residual is: $$
\tilde{Y}_{igt} = Y_{igt} - \hat{\mu}_g - \hat{\eta}_t \xrightarrow{p} (\mu_g + \eta_t + \varepsilon_{igt} + \tau_{igt}) - (\mu_g + \eta_t) = \varepsilon_{igt} + \tau_{igt}
$$

**For treated units:** $\tilde{Y}_{igt} \approx \varepsilon_{igt} + \tau_{igt}$ (idiosyncratic error + treatment effect)

### Cell-Level Averages

#### Untreated Cells ($D_{gt} = 0$)

The average outcome residual in an untreated cell is: $$
\bar{\tilde{Y}}_{gt} = \frac{1}{N_{gt}} \sum_{i=1}^{N_{gt}} \tilde{Y}_{igt} \approx \frac{1}{N_{gt}} \sum_{i=1}^{N_{gt}} \varepsilon_{igt} \xrightarrow{p} 0
$$

**Untreated cell average:** $\bar{\tilde{Y}}_{gt} \approx 0$

#### Treated Cells ($D_{gt} = 1$)

The average outcome residual in a treated cell is: $$
\bar{\tilde{Y}}_{gt} = \frac{1}{N_{gt}} \sum_{i=1}^{N_{gt}} \tilde{Y}_{igt} \approx \frac{1}{N_{gt}} \sum_{i=1}^{N_{gt}} (\varepsilon_{igt} + \tau_{igt})
$$

$$
= \frac{1}{N_{gt}} \sum_{i=1}^{N_{gt}} \varepsilon_{igt} + \frac{1}{N_{gt}} \sum_{i=1}^{N_{gt}} \tau_{igt} \xrightarrow{p} 0 + \bar{\tau}_{gt} = \bar{\tau}_{gt}
$$

Where $\bar{\tau}_{gt} = \frac{1}{N_{gt}} \sum_{i=1}^{N_{gt}} \tau_{igt}$ is the average treatment effect in cell $(g,t)$.

**Treated cell average:** $\bar{\tilde{Y}}_{gt} \approx \bar{\tau}_{gt}$

### Key Insights

#### 1. **Residuals Capture Treatment Effects**

For treated observations, the outcome residuals capture both: - The individual treatment effect $\tau_{igt}$ - The idiosyncratic error $\varepsilon_{igt}$

#### 2. **Cell Averages Isolate Treatment Effects**

When we average within cells, the idiosyncratic errors $\varepsilon_{igt}$ cancel out, leaving: - Untreated cells: $\bar{\tilde{Y}}_{gt} \approx 0$ - Treated cells: $\bar{\tilde{Y}}_{gt} \approx \bar{\tau}_{gt}$

#### 3. **Fixed Effects Provide Counterfactuals**

The estimated fixed effects $\hat{\mu}_g + \hat{\eta}_t$ represent the **counterfactual outcome** that treated units would have had without treatment (their $Y_{igt}^0$).

#### 4. **Parallel Trends is Crucial**

This decomposition only works under parallel trends. If parallel trends fails, then: $$
\hat{\mu}_g + \hat{\eta}_t \not\xrightarrow{p} \mu_g + \eta_t
$$ for treated units, and the residuals won't cleanly separate treatment effects from errors.

## Connection to TWFE Estimation

This explains why the TWFE estimate becomes: $$
\hat{\tau}_{TWFE} = \sum_{g,t} w_{gt} \bar{\tilde{Y}}_{gt} \approx \sum_{g,t: D_{gt}=1} w_{gt} \bar{\tau}_{gt}
$$

The outcome residuals for treated cells directly estimate the cell-specific average treatment effects $\bar{\tau}_{gt}$, which is why TWFE becomes a weighted average of these treatment effects.

The problem is that the weights $w_{gt}$ depend on the problematic treatment residuals $\tilde{D}_{gt}$, not on natural sample-size weighting.

## The TWFE Decomposition

### Starting Point: FWL Second-Stage Regression

Starting with the FWL second-stage regression:

$$
\tilde{y}_{igt} = \tau \tilde{D}_{gt} + \tilde{\varepsilon}_{igt}
$$

The OLS estimator is:

$$
\hat{\tau}_{TWFE} = \frac{\sum_{i,g,t} \tilde{D}_{gt} \tilde{y}_{igt}}{\sum_{i,g,t} \tilde{D}_{gt}^2}
$$

but the residuals $\tilde{y}_{igt}$ are zero when $D_{gt}=0$ and $\bar{\tau}_{gt}$ otherwise, so

$$
\hat{\tau}_{TWFE} = \frac{\sum_{igt:D_{g,t}=1} \tilde{D}_{gt} \bar{\tau}_{gt}}{\sum_{igt} \tilde{D}_{gt}^2}
$$

### Decomposition by Group-Time Cells

We can rewrite this by grouping observations within each $(g,t)$ cell where, in general, we have $N_{gt}$ samples:

$$
\hat{\tau}_{TWFE} = \frac{\sum_{g,t} \tilde{D}_{gt} \sum_{i=1}^{N_{gt}} \tilde{y}_{igt}}{\sum_{g,t} N_{gt} \tilde{D}_{gt}^2}
$$

$$
= \frac{\sum_{g,t} N_{gt} \tilde{D}_{gt} \bar{\tilde{y}}_{gt}}{\sum_{g,t} N_{gt} \tilde{D}_{gt}^2}
$$

Where the average residualized outcome in cell $(g,t)$ is:

$$
\bar{\tilde{y}}_{gt} = \frac{1}{N_{gt}} \sum_{i=1}^{N_{gt}} \tilde{y}_{igt}
$$

### Weighted Average Form

This can be expressed as:

$$
\hat{\tau}_{TWFE} = \sum_{g,t} w_{gt} \bar{\tilde{y}}_{gt}
$$

Where the weights are:

$$
w_{gt} = \frac{N_{gt} \tilde{D}_{gt}}{\sum_{g',t'} N_{g't'} \tilde{D}_{g't'}}
$$

**Note:** These weights can be **negative** when $\tilde{D}_{gt} < 0$, and they sum to 1:

$$
\sum_{g,t} w_{gt} = 1
$$

```{r}
#| echo: false
sum(data$treated_resid * data$outcome_resid) /
  sum(data$treated_resid^2)

data$treated_resid/sum(data$treated_resid^2)

```

```{r}
#| echo: false
fixeff_out <- 
  (outcome_on_fe$fitted.values - outcome_on_fe$fitted.values[1]) |>
    matrix(nrow=2, dimnames = list(c("A","B"),c("1","2")))

fixeff_treat <- 
  (treatment_on_fe$fitted.values - treatment_on_fe$fitted.values[1]) |>
    matrix(nrow=2, dimnames = list(c("A","B"),c("1","2")))

data$treated |>
    matrix(nrow=2, dimnames = list(c("A","B"),c("1","2")))

data$treated - (treatment_on_fe$fitted.values - treatment_on_fe$fitted.values[1])|>
    matrix(nrow=2, dimnames = list(c("A","B"),c("1","2")))

data$outcome - (outcome_on_fe$fitted.values - outcome_on_fe$fitted.values[1])|>
    matrix(nrow=2, dimnames = list(c("A","B"),c("1","2")))

(data$treated - (treatment_on_fe$fitted.values - treatment_on_fe$fitted.values[1]))^2 |> sum()
```

```{r}
#| echo: false
untreated_data <- data[data$treated == 0, ]
cat("Stage 1 data (untreated only):\n")
print(untreated_data[, c("group", "period", "outcome")])

stage1 <- fixest::feols(outcome ~ 1 | group_id + period_id, data = untreated_data)

# Get predicted values for all observations
data$outcome_pred_stage1 <- predict(stage1, newdata = data)
data$outcome_resid_gardner <- data$outcome - data$outcome_pred_stage1

cat("\nStage 1 predictions and residuals:\n")
print(data[, c("group", "period", "outcome", "outcome_pred_stage1", "outcome_resid_gardner")])

# Stage 2: Regress residualized outcome on ORIGINAL treatment indicator
stage2 <- fixest::feols(outcome_resid_gardner ~ treated, data = data)

resid(stage1)

```

```{r}
#| echo: false
data_h3 <- data.frame(
  group = rep(c("A", "B", "C"), each = 3),
  period = rep(1:3, 3),
  group_id = rep(c(1, 2, 3), each = 3),
  period_id = rep(1:3, 3)
) |> 
  dplyr::mutate(
    # Treatment indicator (staggered timing)
    treated = dplyr::case_when(
      group == "A" & period >= 2 ~ 1,
      group == "B" & period >= 3 ~ 1,
      TRUE ~ 0
    )
  )

# Create outcomes with HOMOGENEOUS (constant) treatment effects
data_h3 <- data_h3 |> 
  dplyr::mutate(
    # Base outcome (group and time fixed effects)
    base_outcome = dplyr::case_when(
      group == "A" ~ 10,  # Group A baseline
      group == "B" ~ 12,  # Group B baseline  
      group == "C" ~ 8    # Group C baseline
    ) + (period - 1) * 1.5,  # Common time trend
    
    # HOMOGENEOUS treatment effect: same for all treated units!
    treatment_effect = ifelse(treated == 1, 3.0, 0),  # Constant effect of 3.0
    
    # Final outcome
    outcome = base_outcome + treatment_effect
  )

# estimators
traditional_twfe_h3 <- fixest::feols(outcome ~ treated | group_id + period_id, data = data_h3)

cat("Traditional TWFE Estimate:", round(coef(traditional_twfe_h3)[1], 3), "\n")

# true_effect
true_effects <- data_h3 |>
  dplyr::filter(treated == 1) |>
  dplyr::select(group, period, treatment_effect)

true_att <- mean(true_effects$treatment_effect)

# Analyze the weighting
treatment_on_fe_h3 <- fixest::feols(treated ~ 1 | group_id + period_id, data = data_h3)
data_h3$treated_resid <- resid(treatment_on_fe_h3)

outcome_on_fe_h3 <- fixest::feols(outcome ~ 1 | group_id + period_id, data = data_h3)
data_h3$outcome_resid <- resid(outcome_on_fe_h3)

lm(outcome_resid ~ treated_resid, data = data_h3)

# Calculate implicit weights
nonzero_resid_h3 <- data_h3[abs(data_h3$treated_resid) > 1e-10, ]
if(nrow(nonzero_resid_h3) > 0) {
  weights <- nonzero_resid_h3$treated_resid^2 / sum(nonzero_resid_h3$treated_resid^2)
  
  cat("\nImplicit weights in TWFE estimation:\n")
  weight_table <- data.frame(
    group = nonzero_resid_h3$group,
    period = nonzero_resid_h3$period,
    treated = nonzero_resid_h3$treated,
    treated_resid = round(nonzero_resid_h3$treated_resid, 3),
    weight = round(weights, 3),
    true_effect = nonzero_resid_h3$treatment_effect
  )
  print(weight_table)
  
  # Since all treatment effects are the same, weighted average = unweighted average
  weighted_estimate <- sum(weights * weight_table$true_effect)
  cat("\nWeighted average using TWFE weights:", round(weighted_estimate, 3), "\n")
}

nonzero_resid_h3 |> 
  dplyr::mutate(
    weight = treated_resid / sum(treated_resid^2)
    , comp = outcome_resid * weight)
```

```{r}
#| echo: false
data_nh3 <- data.frame(
  group = rep(c("A", "B", "C"), each = 3),
  period = rep(1:3, 3),
  group_id = rep(c(1, 2, 3), each = 3),
  period_id = rep(1:3, 3)
) |> 
  dplyr::mutate(
    # Treatment indicator (staggered timing)
    treated = dplyr::case_when(
      group == "A" & period >= 2 ~ 1,
      group == "B" & period >= 3 ~ 1,
      TRUE ~ 0
    )
  )

# Create outcomes with HOMOGENEOUS (constant) treatment effects
data_nh3 <- data_nh3 |> 
  dplyr::mutate(
    # Base outcome (group and time fixed effects)
    base_outcome = dplyr::case_when(
      group == "A" ~ 10,  # Group A baseline
      group == "B" ~ 12,  # Group B baseline  
      group == "C" ~ 8    # Group C baseline
    ) + (period - 1) * 1.5,  # Common time trend
    
    # HOMOGENEOUS treatment effect: same for all treated units!
    #treatment_effect = ifelse(treated == 1, 3.0, 0),  # Constant effect of 3.0
    treatment_effect = dplyr::case_when(
      group == "A" & treated == 1 ~ 3,
      group == "B" & treated == 1 ~ 4,
      TRUE ~ 0
    ),
    # Final outcome
    outcome = base_outcome + treatment_effect
  )

# ===========================
# estimators
traditional_twfe_nh3 <- fixest::feols(outcome ~ treated | group_id + period_id, data = data_nh3)

cat("Traditional TWFE Estimate:", round(coef(traditional_twfe_nh3)[1], 3), "\n")

# true_effect
true_effects <- data_nh3 |>
  dplyr::filter(treated == 1) |>
  dplyr::select(group, period, treatment_effect)

true_att <- mean(true_effects$treatment_effect)

# Analyze the weighting
treatment_on_fe_nh3 <- fixest::feols(treated ~ 1 | group_id + period_id, data = data_nh3)
data_nh3$treated_resid <- resid(treatment_on_fe_nh3)

outcome_on_fe_nh3 <- fixest::feols(outcome ~ 1 | group_id + period_id, data = data_nh3)
data_nh3$outcome_resid <- resid(outcome_on_fe_nh3)

# lm(outcome_resid ~ treated_resid, data = data_nh3) |> 
#   broom::tidy() |> dplyr::pull(treated_resid)

# Calculate implicit weights
nonzero_resid_nh3 <- data_nh3[abs(data_nh3$treated_resid) > 1e-10, ]
if(nrow(nonzero_resid_nh3) > 0) {
  weights <- nonzero_resid_nh3$treated_resid^2 / sum(nonzero_resid_nh3$treated_resid^2)
  
  cat("\nImplicit weights in TWFE estimation:\n")
  weight_table <- data.frame(
    group = nonzero_resid_nh3$group,
    period = nonzero_resid_nh3$period,
    treated = nonzero_resid_nh3$treated,
    treated_resid = round(nonzero_resid_nh3$treated_resid, 3),
    weight = round(weights, 3),
    true_effect = nonzero_resid_nh3$treatment_effect
  )
  print(weight_table)
  
  # Since all treatment effects are the same, weighted average = unweighted average
  weighted_estimate <- sum(weights * weight_table$true_effect)
  cat("\nWeighted average using TWFE weights:", round(weighted_estimate, 3), "\n")
}

nonzero_resid_nh3 |> 
  dplyr::mutate(
    weight = treated_resid / sum(treated_resid^2)
    , comp = outcome_resid * weight) |> 
  dplyr::summarize(effect = sum(comp))

# Stage 1: Estimate fixed effects using ONLY UNTREATED observations
untreated_data_nh3 <- data_nh3[data_nh3$treated == 0, ]
stage1 <- fixest::feols(outcome ~ 1 | group_id + period_id, data = untreated_data_nh3)

# Stage 2: Regress residualized outcome on treatment
data_nh3$outcome_resid_gardner <- data_nh3$outcome - predict(stage1, newdata = data_nh3)
stage2 <- fixest::feols(outcome_resid_gardner ~ treated, data = data_nh3)

cat("Gardner Two-Stage Estimate:", round(coef(stage2)[2], 3), "\n")

tibble::tibble(
  "True ATT" = true_att
  , "Traditional TWFE Estimate" = coef(traditional_twfe_nh3)[1]
  , "FWL Estimated" = lm(outcome_resid ~ treated_resid, data = data_nh3) |> 
  broom::tidy() |> dplyr::filter(term == 'treated_resid') |> dplyr::pull(estimate)
  , "Weighted average using TWFE weights" =
    nonzero_resid_nh3 |> 
      dplyr::mutate(
        weight = treated_resid / sum(treated_resid^2)
        , comp = outcome_resid * weight
      ) |> 
      dplyr::summarize(effect = sum(comp)) |> 
      dplyr::pull(effect)
  , "Gardner Two-Stage Estimate" = coef(stage2)[2]
) |> 
  tidyr::pivot_longer(everything()) |> gt::gt()

```

```{r}
#| echo: false
data_nh3$treated_resid |> mean()

denom <- data_nh3 |> dplyr::filter(treated == 1) |> dplyr::pull(treated_resid) |> mean()
unscaled_weights <- data_nh3 |> dplyr::filter(treated == 1) |> dplyr::pull(treated_resid)

weights <- data_nh3 |> dplyr::filter(treated == 1) |> dplyr::pull(treated_resid)/(data_nh3 |> dplyr::filter(treated == 1) |> dplyr::pull(treated_resid) |> mean())

data_nh3$treated_resid |> matrix(ncol = 3)

data_nh3$outcome_resid
```
