---
title: "TwoStageDiD_draft"
format: 
  html:
    self-contained: true
---

::: cell
```{css}
#| echo: false
pre code {
  font-size: 0.7em;
}
.scrolling {
  max-height: 200px;
  overflow-y: auto;
}
.etable {
   width: 400px; /* I took the width from your post and placed it in css */
   height: 400px;
}
/* This will style any <img> element in .parent div */
.etable img {
   height: 80%;
   width: 80%;
}
```
:::

## Starting Assumptions

### The True Data Generating Process

Along with continuous outcomes $y_{igt}$, our data consists of (discrete) units $i$, (discrete) times $t$, and (discrete) groups, indexed by time first treated $g$. Under parallel trends, outcomes follow:

$$
\begin{align}
y_{igt} = \mu_g + \eta_t + \tau_{gt} D_{gt} + \varepsilon_{igt}
\end{align}
$$ {#eq-eq1}

Where:

-   $\mu_g$ = group fixed effect: *effects constant over all time, by group*
-   $\eta_t$ = time fixed effect: *effects constant over all groups, by time*
-   $D_{gt}\equiv \textbf{1}_{g\le t}$: *treatment indicator ([**note**]{.underline} - once treated, always treated)*
-   $\tau_{gt}$ = group-time specific treatment effect: *i.e.* $\mathbb{E}[Y^1_{gt}-Y^0_{gt}]$ *in potential outcomes notation*
-   $\varepsilon_{igt}$ = idiosyncratic error with $E[\varepsilon_{igt}] = 0$

## The simplest case:

$2\times 2$ DiD: two groups, two periods ($t\in\{1,2\}$); one group (A, aka *control* ; $g=\infty$) never treated, the other group (B, aka *treatment* ; $g=2$) treated in the second period. In the data here we have one unit per group, but in practice there will be $N_{gt}$ units in a group-time.

Under the assumptions:

1.  **parallel trends**: that the untreated potential outcomes for both groups have parallel trends.
2.  **no anticipation**: treatment effects are zero in all periods before treatment actually begins.

then $\tau_{gt}$ represents a causal treatment effect (the estimand).

In this case, a DiD estimator is "four means and three subtractions" (the reason this method is called diff-in-diff).

```{r}
#| echo: true
#| code-fold: true
#| code-summary: classic 2x2 example data is
#| label: classic 2x2 example data is
# Create the classic 2x2 difference-in-differences setup
data <- data.frame(
  group = c("control", "control", "treatment", "treatment"),
  period = c(1, 2, 1, 2),
  treated = c(0, 0, 0, 1),
  # Create outcomes with known treatment effect
  outcome = c(10, 11, 12, 16)  # Treatment effect = 3
)

data$group_id <- ifelse(data$group == "control", 1, 2)
data$period_id <- data$period

data |> gt::gt() |> gtExtras::gt_theme_espn()
```

with matrix representation:

```{r}
#| echo: true
#| code-fold: true
#| code-summary: classic 2x2 data matrix
#| layout-align: center
#| label: classic 2x2 data matrix
data$outcome |> 
  matrix(nrow=2, dimnames = list(c("A","B"),c("1","2")))
```

This estimator in general is calculated by taking the means of each group, subtracting the *pre* means from the *post* means of each group, and then subtracting the treated differences from the untreated differences:

```{r}
#| echo: true
#| code-fold: true
#| code-summary: classic 2x2 four means and three subtractions
#| label: classic 2x2 four means and three subtractions
# Classic DiD calculation: 
# (Y_treat_post - Y_treat_pre) - (Y_control_post - Y_control_pre)
true_effect <- (16 - 12) - (11 - 10)
cat("\nTRUE Treatment Effect (manual DiD):", true_effect, "\n\n")
```

The fixed effects regression (@eq-eq1) gives the same estimate (The `fixest` package offers a family of functions to perform estimations with multiple fixed-effects):

```{r}
#| echo: true
#| code-fold: true
#| code-summary: classic 2x2 by TWFE
#| label: classic 2x2 by TWFE

traditional_twfe <- fixest::feols(outcome ~ treated | group_id + period_id, data = data)

cat("Traditional TWFE Estimate:", coef(traditional_twfe)["treated"], "\n")
```

Finally, we use the Frisch-Waugh-Lovell (FWL) Theorem to estimate the TWFE model.

Per the FWL Theorem, the TWFE model is equivalent to the residualized regression: $y_{igt} - \hat{\mu}_g - \hat{\eta}_t = \tau \tilde{D}_{gt} + \tilde{\varepsilon}_{gt}$

1.  starting with $y_{igt} = \mu_g + \eta_t + \tau_{gt}D_{gt} + \varepsilon_{igt}$ regress the output $y_{igt}$ on fixed effects (say $[\mu_g, \eta_t]$)
    -   the fitted values are $\hat{y}_{igt} = \hat{\mu}_g + \hat{\eta}_t$ and so the residuals are $\tilde{y}_{igt} = y_{igt} - \hat{\mu}_g -\hat{\eta}_t$
2.  regress the treatment indicator $D_{gt}$ on fixed effects (say $[\alpha_g, \delta_t]$)
    -   the fitted values are $\hat{D}_{gt}$ and so the residuals are $\tilde{D}_{gt} = D_{gt} -\hat{D}_{gt}$
3.  regress the output residuals on the treatment residuals $\tilde{y}_{igt} = \tau\tilde{D}_{gt} + \tilde{\varepsilon}_{gt}$

```{r}
#| echo: true
#| code-fold: true
#| code-summary: Step 1 - Regress outcome on fixed effects
#| label: Step 1 - Regress outcome on fixed effects
outcome_on_fe <- 
  fixest::feols(outcome ~ 1 | group_id + period_id, data = data)
data$outcome_resid <- resid(outcome_on_fe)
```

:::::: columns
::: {.column width="45%"}
```{r}
#| echo: true
#| code-fold: true
#| code-summary: Step 1 - outcome fitted values
#| label: Step 1 - outcome fitted values
outcome_on_fe$fitted.values |>
  matrix(nrow=2, dimnames = list(c("A","B"),c("1","2")))
```
:::

::: {.column width="10%"}
<!-- empty column to create gap -->
:::

::: {.column width="45%"}
```{r}
#| echo: true
#| code-fold: true
#| code-summary: Step 1 - outcome fitted residuals
#| label: Step 1 - outcome fitted residuals
resid(outcome_on_fe) |> 
  matrix(nrow=2, dimnames = list(c("A","B"),c("1","2")))
```
:::
::::::

We can now compute the treatment residuals the same way and then regress the outcome residuals on the treatment residuals, per FWL

```{r}
#| echo: true
#| code-fold: true
#| code-summary: Step 2 - Regress outcome residuals on treatment residuals
#| label: Step 2 - Regress outcome residuals on treatment residuals
#| warning: false
# regress treatment on fixed effects
treatment_on_fe <- 
  fixest::feols(treated ~ 1 | group_id + period_id, data = data)
data$treated_resid <- resid(treatment_on_fe)

# use residuals to estimate TWFE regression
lm(outcome_resid ~ treated_resid, data = data) |> broom::tidy()
```

And we see that the calculation per FWL is the same as computed by the other two estimators. Note that the FWL estimate can be calculated directly using the residuals

$$
\beta_{FWL}=\frac{\text{cov}(\tilde{D}_{gt},\tilde{y}_{igt})}{\text{var}(\tilde{D}_{gt})}
$$

```{r}
#| echo: true
#| code-fold: true
#| code-summary: variance-covariance calculation
#| label: Step 2 - variance-covariance calculation
cov(data$treated_resid, data$outcome_resid) /
  var(data$treated_resid)
```

This is equal to the alternative formulation using the mean of the outcomes and the treatment indicator.

$$
\beta_{FWL}=\frac{\text{cov}(\tilde{D}_{gt},\hat{y}_{igt})}{\text{cov}(\tilde{D}_{gt},D_{gt})}
$$

```{r}
#| echo: true
#| code-fold: true
#| code-summary: alternate variance-covariance calculation
#| label: Step 2 - alternate variance-covariance calculation
cov(data$treated_resid, data$outcome_resid) /
  var(data$treated_resid)
cov(data$treated_resid, data$outcome) /
  cov(data$treated_resid, data$treated)
```

We'll now look a little deeper at how the result was arrived at with the FWL estimator, and what it exposes about the structure of the treatment estimator.

## Outcome Residuals in Terms of Potential Outcomes

### Setup: The Data Generating Process

Under the potential outcomes framework, each unit has two potential outcomes: (i) $Y_{igt}^0$ = potential outcome under no treatment, and (ii) $Y_{igt}^1$ = potential outcome under treatment

The observed outcome is: $$
Y_{igt} = Y_{igt}^0 + D_{gt}(Y_{igt}^1 - Y_{igt}^0) = Y_{igt}^0 + D_{gt}\tau_{igt}
$$

Where $\tau_{igt} = Y_{igt}^1 - Y_{igt}^0$ is the individual treatment effect.

### Parallel Trends Assumption

Under parallel trends, the untreated potential outcomes follow: $$
Y_{igt}^0 = \mu_g + \eta_t + \varepsilon_{igt}
$$

Where: $\mu_g$ = group fixed effect (time-invariant group characteristics), $\eta_t$ = time fixed effect (group-invariant time shocks), and $\varepsilon_{igt}$ = idiosyncratic error with $E[\varepsilon_{igt}] = 0$

### The FWL First Stage: Outcome on Fixed Effects

When we regress outcomes on fixed effects using **all observations**: $$
Y_{igt} = \hat{\mu}_g + \hat{\eta}_t + \text{residual}_{igt}
$$

The outcome residuals are: $$
\tilde{Y}_{igt} = Y_{igt} - \hat{\mu}_g - \hat{\eta}_t
$$

### Decomposition by Treatment Status

#### Case 1: Untreated Observations ($D_{gt} = 0$)

For untreated units, the observed outcome is: $$
Y_{igt} = Y_{igt}^0 = \mu_g + \eta_t + \varepsilon_{igt}
$$

Under consistent fixed effects estimation: $$
\hat{\mu}_g \xrightarrow{p} \mu_g \quad \text{and} \quad \hat{\eta}_t \xrightarrow{p} \eta_t
$$

Therefore, the outcome residual is: $$
\tilde{Y}_{igt} = Y_{igt} - \hat{\mu}_g - \hat{\eta}_t \xrightarrow{p} (\mu_g + \eta_t + \varepsilon_{igt}) - (\mu_g + \eta_t) = \varepsilon_{igt}
$$

**For untreated units:** $\tilde{Y}_{igt} \approx \varepsilon_{igt}$ (idiosyncratic error)

#### Case 2: Treated Observations ($D_{gt} = 1$)

For treated units, the observed outcome is: $$
Y_{igt} = Y_{igt}^1 = Y_{igt}^0 + \tau_{igt} = \mu_g + \eta_t + \varepsilon_{igt} + \tau_{igt}
$$

The crucial insight is that $\hat{\mu}_g$ and $\hat{\eta}_t$ are estimated from **both treated and untreated observations**. Under parallel trends, these estimates converge to the **counterfactual** fixed effects: $$
\hat{\mu}_g \xrightarrow{p} \mu_g \quad \text{and} \quad \hat{\eta}_t \xrightarrow{p} \eta_t
$$

Therefore, the outcome residual is: $$
\tilde{Y}_{igt} = Y_{igt} - \hat{\mu}_g - \hat{\eta}_t \xrightarrow{p} (\mu_g + \eta_t + \varepsilon_{igt} + \tau_{igt}) - (\mu_g + \eta_t) = \varepsilon_{igt} + \tau_{igt}
$$

**For treated units:** $\tilde{Y}_{igt} \approx \varepsilon_{igt} + \tau_{igt}$ (idiosyncratic error + treatment effect)

### Cell-Level Averages

#### Untreated Cells ($D_{gt} = 0$)

The average outcome residual in an untreated cell is: $$
\bar{\tilde{Y}}_{gt} = \frac{1}{N_{gt}} \sum_{i=1}^{N_{gt}} \tilde{Y}_{igt} \approx \frac{1}{N_{gt}} \sum_{i=1}^{N_{gt}} \varepsilon_{igt} \xrightarrow{p} 0
$$

**Untreated cell average:** $\bar{\tilde{Y}}_{gt} \approx 0$

#### Treated Cells ($D_{gt} = 1$)

The average outcome residual in a treated cell is:

$$
\bar{\tilde{Y}}_{gt} = \frac{1}{N_{gt}} \sum_{i=1}^{N_{gt}} \tilde{Y}_{igt} \approx \frac{1}{N_{gt}} \sum_{i=1}^{N_{gt}} (\varepsilon_{igt} + \tau_{igt})
$$

$$
= \frac{1}{N_{gt}} \sum_{i=1}^{N_{gt}} \varepsilon_{igt} + \frac{1}{N_{gt}} \sum_{i=1}^{N_{gt}} \tau_{igt} \xrightarrow{p} 0 + \bar{\tau}_{gt} = \bar{\tau}_{gt}
$$

Where $\bar{\tau}_{gt} = \frac{1}{N_{gt}} \sum_{i=1}^{N_{gt}} \tau_{igt}$ is the average treatment effect in cell $(g,t)$.

**Treated cell average:** $\bar{\tilde{Y}}_{gt} \approx \bar{\tau}_{gt}$

### Key Insights

#### 1. **Residuals Capture Treatment Effects**

For treated observations, the outcome residuals capture both: (i) the individual treatment effect $\tau_{igt}$, and (ii) The idiosyncratic error $\varepsilon_{igt}$.

#### 2. **Cell Averages Isolate Treatment Effects**

When we average within cells, the idiosyncratic errors $\varepsilon_{igt}$ cancel out, leaving: - Untreated cells: $\bar{\tilde{Y}}_{gt} \approx 0$ - Treated cells: $\bar{\tilde{Y}}_{gt} \approx \bar{\tau}_{gt}$.

#### 3. **Fixed Effects Provide Counterfactuals**

The estimated fixed effects $\hat{\mu}_g + \hat{\eta}_t$ represent the **counterfactual outcome** that treated units would have had without treatment (their $Y_{igt}^0$).

#### 4. **Parallel Trends is Crucial**

This decomposition only works under parallel trends. If parallel trends fails, then:

$$
\hat{\mu}_g + \hat{\eta}_t \not\xrightarrow{p} \mu_g + \eta_t
$$ for treated units, and the residuals won't cleanly separate treatment effects from errors.

## Connection to TWFE Estimation

This explains why the TWFE estimate becomes: $$
\hat{\tau}_{TWFE} = \sum_{g,t} w_{gt} \bar{\tilde{Y}}_{gt} \approx \sum_{g,t: D_{gt}=1} w_{gt} \bar{\tau}_{gt}
$$

The outcome residuals for treated cells directly estimate the cell-specific average treatment effects $\bar{\tau}_{gt}$, which is why TWFE becomes a weighted average of these treatment effects.

The problem is that the weights $w_{gt}$ depend on the problematic treatment residuals $\tilde{D}_{gt}$, not on natural sample-size weighting.

## The TWFE Decomposition

### Starting Point: FWL Second-Stage Regression

Starting with the FWL second-stage regression:

$$
\tilde{y}_{igt} = \tau_{TWFE} \times\tilde{D}_{gt} + \tilde{\varepsilon}_{igt}
$$

The OLS estimator is:

$$
\hat{\tau}_{TWFE} =\frac{\sum_{i,g,t}\tilde{D}_{gt}y_{igt}}{\sum_{i,g,t}\tilde{D}_{gt}D_{gt}}
$$

We can rewrite this by grouping observations within each $(g,t)$ cell where, in general, we have $N_{gt}$ samples:

$$
\begin{align*}
\hat{\tau}_{TWFE} & =\frac{\sum_{i,g,t}\tilde{D}_{gt}y_{igt}}{\sum_{i,g,t}\tilde{D}_{gt}D_{gt}}\\
 & =\frac{\sum_{g,t}N_{gt}\tilde{D}_{gt}\frac{1}{N_{gt}}\sum_{i=1}^{N_{gt}}y_{igt}}{\sum_{g,t}N_{gt}\tilde{D}_{gt}\frac{1}{N_{gt}}\sum_{i=1}^{N_{gt}}D_{gt}}\\
 & =\frac{\sum_{gt:D_{g,t}=1}N_{gt}\tilde{D}_{gt}\bar{\tau}_{gt}}{\sum_{gt:D_{g,t}=1}N_{gt}\tilde{D}_{gt}}\\
 & =\sum_{gt:D_{g,t}=1}\frac{N_{gt}}{N_{1}}w_{gt}\tau_{gt}
\end{align*}
$$

### Weighted Average Form

This can be expressed as:

$$
\hat{\tau}_{TWFE} = \sum_{gt:D_{g,t}=1}\frac{N_{gt}}{N_{1}}w_{gt}\tau_{gt}
$$

Where the weights are:

$$
w_{gt}=\frac{\tilde{D}_{gt}}{\sum_{gt:D_{g,t}=1}N_{gt}\tilde{D}_{gt}}
$$

**Note:** These weights can be **negative** when $\tilde{D}_{gt} < 0$, and they sum to 1:

$$
\sum_{g,t} w_{gt} = 1
$$

```{r}
#| echo: false
#| eval: false
#| label: misc chunk0

sum(data$treated_resid * data$outcome_resid) /
  sum(data$treated_resid^2)

data$treated_resid/sum(data$treated_resid^2)

```

```{r}
#| echo: false
#| eval: false
#| label: misc chunk1

fixeff_out <- 
  (outcome_on_fe$fitted.values - outcome_on_fe$fitted.values[1]) |>
    matrix(nrow=2, dimnames = list(c("A","B"),c("1","2")))

fixeff_treat <- 
  (treatment_on_fe$fitted.values - treatment_on_fe$fitted.values[1]) |>
    matrix(nrow=2, dimnames = list(c("A","B"),c("1","2")))

data$treated |>
    matrix(nrow=2, dimnames = list(c("A","B"),c("1","2")))

data$treated - (treatment_on_fe$fitted.values - treatment_on_fe$fitted.values[1])|>
    matrix(nrow=2, dimnames = list(c("A","B"),c("1","2")))

data$outcome - (outcome_on_fe$fitted.values - outcome_on_fe$fitted.values[1])|>
    matrix(nrow=2, dimnames = list(c("A","B"),c("1","2")))

(data$treated - (treatment_on_fe$fitted.values - treatment_on_fe$fitted.values[1]))^2 |> sum()
```

```{r}
#| echo: false
#| eval: false
#| label: misc chunk2
untreated_data <- data[data$treated == 0, ]
cat("Stage 1 data (untreated only):\n")
print(untreated_data[, c("group", "period", "outcome")])

stage1 <- fixest::feols(outcome ~ 1 | group_id + period_id, data = untreated_data)

# Get predicted values for all observations
data$outcome_pred_stage1 <- predict(stage1, newdata = data)
data$outcome_resid_gardner <- data$outcome - data$outcome_pred_stage1

cat("\nStage 1 predictions and residuals:\n")
print(data[, c("group", "period", "outcome", "outcome_pred_stage1", "outcome_resid_gardner")])

# Stage 2: Regress residualized outcome on ORIGINAL treatment indicator
stage2 <- fixest::feols(outcome_resid_gardner ~ treated, data = data)

resid(stage1)

```

```{r}
#| echo: false
#| eval: false
#| label: misc chunk3
data_h3 <- data.frame(
  group = rep(c("A", "B", "C"), each = 3),
  period = rep(1:3, 3),
  group_id = rep(c(1, 2, 3), each = 3),
  period_id = rep(1:3, 3)
) |> 
  dplyr::mutate(
    # Treatment indicator (staggered timing)
    treated = dplyr::case_when(
      group == "A" & period >= 2 ~ 1,
      group == "B" & period >= 3 ~ 1,
      TRUE ~ 0
    )
  )

# Create outcomes with HOMOGENEOUS (constant) treatment effects
data_h3 <- data_h3 |> 
  dplyr::mutate(
    # Base outcome (group and time fixed effects)
    base_outcome = dplyr::case_when(
      group == "A" ~ 10,  # Group A baseline
      group == "B" ~ 12,  # Group B baseline  
      group == "C" ~ 8    # Group C baseline
    ) + (period - 1) * 1.5,  # Common time trend
    
    # HOMOGENEOUS treatment effect: same for all treated units!
    treatment_effect = ifelse(treated == 1, 3.0, 0),  # Constant effect of 3.0
    
    # Final outcome
    outcome = base_outcome + treatment_effect
  )

# estimators
traditional_twfe_h3 <- fixest::feols(outcome ~ treated | group_id + period_id, data = data_h3)

cat("Traditional TWFE Estimate:", round(coef(traditional_twfe_h3)[1], 3), "\n")

# true_effect
true_effects <- data_h3 |>
  dplyr::filter(treated == 1) |>
  dplyr::select(group, period, treatment_effect)

true_att <- mean(true_effects$treatment_effect)

# Analyze the weighting
treatment_on_fe_h3 <- fixest::feols(treated ~ 1 | group_id + period_id, data = data_h3)
data_h3$treated_resid <- resid(treatment_on_fe_h3)

outcome_on_fe_h3 <- fixest::feols(outcome ~ 1 | group_id + period_id, data = data_h3)
data_h3$outcome_resid <- resid(outcome_on_fe_h3)

lm(outcome_resid ~ treated_resid, data = data_h3)

# Calculate implicit weights
nonzero_resid_h3 <- data_h3[abs(data_h3$treated_resid) > 1e-10, ]
if(nrow(nonzero_resid_h3) > 0) {
  weights <- nonzero_resid_h3$treated_resid^2 / sum(nonzero_resid_h3$treated_resid^2)
  
  cat("\nImplicit weights in TWFE estimation:\n")
  weight_table <- data.frame(
    group = nonzero_resid_h3$group,
    period = nonzero_resid_h3$period,
    treated = nonzero_resid_h3$treated,
    treated_resid = round(nonzero_resid_h3$treated_resid, 3),
    weight = round(weights, 3),
    true_effect = nonzero_resid_h3$treatment_effect
  )
  print(weight_table)
  
  # Since all treatment effects are the same, weighted average = unweighted average
  weighted_estimate <- sum(weights * weight_table$true_effect)
  cat("\nWeighted average using TWFE weights:", round(weighted_estimate, 3), "\n")
}

nonzero_resid_h3 |> 
  dplyr::mutate(
    weight = treated_resid / sum(treated_resid^2)
    , comp = outcome_resid * weight)
```

```{r}
#| echo: false
#| eval: false
#| label: misc chunk4
data_nh3 <- data.frame(
  group = rep(c("A", "B", "C"), each = 3),
  period = rep(1:3, 3),
  group_id = rep(c(1, 2, 3), each = 3),
  period_id = rep(1:3, 3)
) |> 
  dplyr::mutate(
    # Treatment indicator (staggered timing)
    treated = dplyr::case_when(
      group == "A" & period >= 2 ~ 1,
      group == "B" & period >= 3 ~ 1,
      TRUE ~ 0
    )
  )

# Create outcomes with HOMOGENEOUS (constant) treatment effects
data_nh3 <- data_nh3 |> 
  dplyr::mutate(
    # Base outcome (group and time fixed effects)
    base_outcome = dplyr::case_when(
      group == "A" ~ 10,  # Group A baseline
      group == "B" ~ 12,  # Group B baseline  
      group == "C" ~ 8    # Group C baseline
    ) + (period - 1) * 1.5,  # Common time trend
    
    # HOMOGENEOUS treatment effect: same for all treated units!
    #treatment_effect = ifelse(treated == 1, 3.0, 0),  # Constant effect of 3.0
    treatment_effect = dplyr::case_when(
      group == "A" & treated == 1 ~ 3,
      group == "B" & treated == 1 ~ 4,
      TRUE ~ 0
    ),
    # Final outcome
    outcome = base_outcome + treatment_effect
  )

# ===========================
# estimators
traditional_twfe_nh3 <- fixest::feols(outcome ~ treated | group_id + period_id, data = data_nh3)

cat("Traditional TWFE Estimate:", round(coef(traditional_twfe_nh3)[1], 3), "\n")

# Analyze the weighting
treatment_on_fe_nh3 <- fixest::feols(treated ~ 1 | group_id + period_id, data = data_nh3)
data_nh3$treated_resid <- resid(treatment_on_fe_nh3)

outcome_on_fe_nh3 <- fixest::feols(outcome ~ 1 | group_id + period_id, data = data_nh3)
data_nh3$outcome_resid <- resid(outcome_on_fe_nh3)

# lm(outcome_resid ~ treated_resid, data = data_nh3) |> 
#   broom::tidy() |> dplyr::pull(treated_resid)

# Calculate implicit weights
nonzero_resid_nh3 <- data_nh3[abs(data_nh3$treated_resid) > 1e-10, ]
if(nrow(nonzero_resid_nh3) > 0) {
  weights <- nonzero_resid_nh3$treated_resid^2 / sum(nonzero_resid_nh3$treated_resid^2)
  
  cat("\nImplicit weights in TWFE estimation:\n")
  weight_table <- data.frame(
    group = nonzero_resid_nh3$group,
    period = nonzero_resid_nh3$period,
    treated = nonzero_resid_nh3$treated,
    treated_resid = round(nonzero_resid_nh3$treated_resid, 3),
    weight = round(weights, 3),
    true_effect = nonzero_resid_nh3$treatment_effect
  )
  print(weight_table)
  
  # Since all treatment effects are the same, weighted average = unweighted average
  weighted_estimate <- sum(weights * weight_table$true_effect)
  cat("\nWeighted average using TWFE weights:", round(weighted_estimate, 3), "\n")
}

nonzero_resid_nh3 |> 
  dplyr::mutate(
    weight = treated_resid / sum(treated_resid^2)
    , comp = outcome_resid * weight) |> 
  dplyr::summarize(effect = sum(comp))

# Stage 1: Estimate fixed effects using ONLY UNTREATED observations
untreated_data_nh3 <- data_nh3[data_nh3$treated == 0, ]
stage1 <- fixest::feols(outcome ~ 1 | group_id + period_id, data = untreated_data_nh3)

# Stage 2: Regress residualized outcome on treatment
data_nh3$outcome_resid_gardner <- data_nh3$outcome - predict(stage1, newdata = data_nh3)
stage2 <- fixest::feols(outcome_resid_gardner ~ treated, data = data_nh3)

cat("Gardner Two-Stage Estimate:", round(coef(stage2)[2], 3), "\n")

tibble::tibble(
  "True ATT" = true_att
  , "Traditional TWFE Estimate" = coef(traditional_twfe_nh3)[1]
  , "FWL Estimated" = lm(outcome_resid ~ treated_resid, data = data_nh3) |> 
  broom::tidy() |> dplyr::filter(term == 'treated_resid') |> dplyr::pull(estimate)
  , "Weighted average using TWFE weights" =
    nonzero_resid_nh3 |> 
      dplyr::mutate(
        weight = treated_resid / sum(treated_resid^2)
        , comp = outcome_resid * weight
      ) |> 
      dplyr::summarize(effect = sum(comp)) |> 
      dplyr::pull(effect)
  , "Gardner Two-Stage Estimate" = coef(stage2)[2]
) |> 
  tidyr::pivot_longer(everything()) |> gt::gt("name")

```

```{r}
#| echo: false
#| eval: false
#| label: misc chunk5
data_nh3$treated_resid |> mean()

denom <- data_nh3 |> dplyr::filter(treated == 1) |> dplyr::pull(treated_resid) |> mean()
unscaled_weights <- data_nh3 |> dplyr::filter(treated == 1) |> dplyr::pull(treated_resid)

weights <- data_nh3 |> dplyr::filter(treated == 1) |> dplyr::pull(treated_resid)/(data_nh3 |> dplyr::filter(treated == 1) |> dplyr::pull(treated_resid) |> mean())

data_nh3$outcome |> matrix(ncol = 3,byrow =T)
data_nh3$treated |> matrix(ncol = 3,byrow =T)

data_nh3$treated_resid |> matrix(ncol = 3,byrow =T)

data_nh3$outcome_resid |> matrix(ncol = 3,byrow =T)



denom <- 
  (data_nh3 |> dplyr::filter(treated == 1) |> dplyr::pull(treated_resid)) |> sum()

unscaled_weights <- data_nh3 |> dplyr::filter(treated == 1) |> dplyr::pull(treated_resid)

weights <- unscaled_weights/denom

treated_resid <- 
  data_nh3 |> dplyr::filter(treated == 1) |> dplyr::pull(treated_resid)

treated_eff <- 
  data_nh3 |> dplyr::filter(treated == 1) |> dplyr::pull(treatment_effect)

treated_eff %*% weights


                                                                  
                                                                  

var(data_nh3$treated_resid)
# version 1
cov(data_nh3$treated_resid, data_nh3$outcome_resid)
cov(data_nh3$treated_resid, data_nh3$outcome_resid) / var(data_nh3$treated_resid)

# version 2
cov(data_nh3$treated_resid, data_nh3$outcome)
cov(data_nh3$treated_resid, data_nh3$outcome)/cov(data_nh3$treated_resid,data_nh3$treated)

# Stage 1: Estimate fixed effects using ONLY UNTREATED observations
untreated_data_nh3 <- data_nh3[data_nh3$treated == 0, ]
stage1 <- fixest::feols(outcome ~ 1 | group_id + period_id, data = untreated_data_nh3)

# Stage 2: Regress residualized outcome on treatment
data_nh3$outcome_resid_gardner <- data_nh3$outcome - predict(stage1, newdata = data_nh3)
stage2 <- fixest::feols(outcome_resid_gardner ~ treated, data = data_nh3)

cat("Gardner Two-Stage Estimate:", round(coef(stage2)[2], 3), "\n")

```

## Negative treatment residuals

## Goodman-Bacon decomposition

The introduction of differential treatment timing makes it hard to draw a bright line between *pre* and *post* treatment periods.

To show this we use the following data:

```{r}
#| class-output: .scrolling
#| echo: true
#| code-fold: true
#| code-summary: new data - 3 groups x 10 periods
#| label: new data - 3 groups x 10 periods
dat4_rev = data.frame(
  group_id = rep(1:3, times = 10),
  period_id = rep(1:10, each = 3)
) |>
  within({
    treated = (group_id == 2 & period_id >= 5) | (group_id == 3 & period_id >= 8)
    btrue = ifelse(treated & group_id == 3, 4, ifelse(treated & group_id == 2, 2, 0))
    outcome = group_id + 1 * period_id + btrue * treated
  }) |> 
    dplyr::mutate(
    treated = as.numeric(treated)
    , treatment_effect = 
      dplyr::case_when(
        group_id == 2 & period_id >= 5 ~ 2
        , group_id == 3 & period_id >= 8 ~ 4
        , TRUE ~ 0
      )
  )
dat4_rev
```

```{r}
#| echo: true
#| code-fold: true
#| code-summary: new data plotted
#| label: new data plotted
require(ggplot2, quietly = TRUE)
#theme_set(theme_bw(base_size = 22))
ggplot(dat4_rev, aes(x = period_id, y = outcome, col = factor(group_id))) +
  geom_point() + geom_line() +
  geom_vline(xintercept = c(4.5, 7.5), lty = 2) +
  scale_x_continuous(breaks = scales::pretty_breaks()) +
  labs(x = "Time period", y = "Outcome variable", col = "ID") +
  theme_light(base_size = 18) 
```

Here we see that our simulation includes two distinct treatment periods. The first treatment occurs at period 5, where group_id=2’s trendline jumps by 2 units. The second treatment occurs in period 8, where group_id=3’s trendline jumps by 4 units. In contrast, id=1 remains untreated for the duration of the experiment.

It’s not immediately clear how to calculate the ATT. For example, how should the late treated unit (group_id=3) regard the early treated unit (group_id=2)? Can the latter be used as control group for the former? After all, they didn’t receive treatment at the same time… but, on the other hand, group_id=2’s path was already altered by the initial treatment wave.

Start by estimating a simple TWFE model (three ways):

```{r}
#| echo: true
#| code-fold: true
#| code-summary: TWFE estimated with new data
#| label: TWFE estimated with new data

# (1) using fixest::feols directly
traditional_twfe4_rev <- fixest::feols(outcome ~ treated | group_id + period_id, data = dat4_rev)

# (2) using straight FWL
treatment_on_fe4_rev <- fixest::feols(treated ~ 1 | group_id + period_id, data = dat4_rev)
dat4_rev$treated_resid <- resid(treatment_on_fe4_rev)

outcome_on_fe4_rev <- fixest::feols(outcome ~ 1 | group_id + period_id, data = dat4_rev)
dat4_rev$outcome_resid <- resid(outcome_on_fe4_rev)

treated <- dat4_rev |> dplyr::filter(treated == 1)
# (3) using weights
weights <- treated$treated_resid/sum(treated$treated_resid)

# (3) calculate ATT directly
treated <- dat4_rev |>
  dplyr::filter(treated == 1)
true_ATT <- treated |>
  dplyr::pull(treatment_effect) |> mean()

tibble::tibble(
  "True ATT" = true_ATT
  , "Traditional TWFE Estimate" = coef(traditional_twfe4_rev)[1]
  , "FWL Estimated" = lm(outcome_resid ~ treated_resid, data = dat4_rev) |> 
  broom::tidy() |> dplyr::filter(term == 'treated_resid') |> dplyr::pull(estimate)
  , "Weighted average using TWFE weights" =
     treated$treatment_effect %*% weights
) |> 
  tidyr::pivot_longer(everything()) |> gt::gt('name')
```

So all our methods agree to 6 digits, and they are all wrong. What is going on?

The short answer is that it comprises a *weighted average* of four distinct 2x2 groups (or comparisons):

1.  **treated** vs **untreated** 1) *early treated (*TeTe) vs *untreated (*UU) 2) *late treated (*TlTl) vs *untreated (*UU)

2.  **differentially treated** 1) *early treated (*TeTe) vs *late control (*ClCl) 2) *late treated (*TlTl) vs *early control (*CeCe)

We can visualize these four comparison sets as follows:

```{r}
#| echo: true
#| code-fold: true
#| code-summary: Goodman-Bacon decomposition plot
#| label: Goodman-Bacon decomposition plot
rbind(
  dat4_rev |> subset(group_id %in% c(1,2)) |> 
    transform(role = ifelse(group_id==2, "Treatment", "Control"), comp = "1.1. Early vs Untreated"),
  dat4_rev |> subset(group_id %in% c(1,3)) |> 
    transform(role = ifelse(group_id==3, "Treatment", "Control"), comp = "1.2. Late vs Untreated"),
  dat4_rev |> subset(group_id %in% c(2,3) & period_id<8) |> 
    transform(role = ifelse(group_id==2, "Treatment", "Control"), comp = "2.1. Early vs Untreated"),
  dat4_rev |> subset(group_id %in% c(2:3) & period_id>4) |> 
    transform(role = ifelse(group_id==3, "Treatment", "Control"), comp = "2.2. Late vs Untreated")
) |>
  ggplot(aes(period_id, outcome, group = group_id, col = factor(group_id), lty = role)) +
  geom_point() + geom_line() +
  facet_wrap(~comp) +
  scale_x_continuous(breaks = scales::pretty_breaks()) +
  scale_linetype_manual(values = c("Control" = 5, "Treatment" = 1)) +
  labs(x = "Time period", y = "Outcome variable", col = "ID", lty = "Role") +
  theme_light(base_size = 14) 
```

In other words, the panel group_time IDs are split into different timing cohorts based on when the first treatment takes place and where it lies in relation to the treatment of other panel IDs. The more panel IDs and differential treatment timings there are, the more the combinations of the above groups.

The Goodman-Bacon decomposition reveals a fundamental issue with two-way fixed effects (TWFE) difference-in-differences estimators when treatment timing varies across units: the overall treatment effect estimate is a weighted average of all possible 2×2 DD comparisons, including problematic ones where already-treated units serve as controls for newly-treated units.

The key insight is that TWFE doesn't just compare treated to never-treated units. When treatment rolls out over time, it also compares "early-treated" to "later-treated" groups. This becomes problematic because already-treated units are poor controls—they've already experienced treatment effects, potentially violating the parallel trends assumption for the comparison group.

More troublingly, some of these comparisons can receive *negative* weights in the overall average. This means that if treatment effects grow over time (dynamic effects are present), or if effects are heterogeneous across groups, TWFE can produce severely biased estimates—even getting the wrong sign. For instance, a beneficial treatment might appear harmful because the negative-weighted comparisons dominate.

The decomposition essentially unpacks the "black box" of TWFE, showing researchers exactly which comparisons drive their estimates and with what weights. This exposed why many DD studies with staggered adoption were likely misestimating treatment effects, spurring the development of new estimators (like Callaway-Sant'Anna, Sun-Abraham, and others) that avoid these problematic comparisons by only using not-yet-treated or never-treated units as controls.

The practical implication: researchers can no longer naively apply TWFE to staggered adoption designs without carefully considering treatment effect dynamics and heterogeneity.

## The right approach

How to explain the failure of TWFE regression to compute the correct ATT when we have differential treatment timing or heterogeneous effects? One explanation is that the treated values bias the computation of the fixed effects.

Looking at the group-time outcomes $y_{gt}$ (since we have just one unit of data per group-time):

```{r}
#| echo: true
#| code-fold: true
#| code-summary: show data by group-time
#| label: show data by group-time
#| fig-cap: Outcome data by group-time
dmnms <- 
  list(
    paste("group",as.character(1:3))
    , as.character(1:10)
    )
dat4_rev |> dplyr::arrange(group_id) |> 
  dplyr::pull(outcome) |> round(2) |> 
  matrix(nrow = 3,byrow =T, dimnames = dmnms) |> 
  as.data.frame() |> 
  knitr::kable("html",digits=2,table.attr = "style='width:40%;'") |> 
  kableExtra::column_spec(column = 1, width = "5em") |> 
  kableExtra::column_spec(column = 2:11, width = "3em") |> 
  kableExtra::kable_styling(full_width = F) |> 
  kableExtra::footnote(footnote_as_chunk=T, general = "group 1 is never treated.")
```

By definition fixed effects are constant by group and by time. By convention, the group effects are zero for group-times (1,.) and the time effects are zero for group-times (.,1). Regressing the output against the fixed effects, should estimate the fixed effect, and the residual outcomes should then just be the causal effects, but this does not happen if we use all the outcomes in the regression:

```{r}
#| echo: true
#| code-fold: true
#| code-summary: show residuals by group-time using all outcome data
#| label: show residuals by group-time using all outcome data
#| fig-cap: Residual outcome data by group-time, estimates using all outomes
dat4_rev |> dplyr::arrange(group_id) |> 
  dplyr::pull(outcome_resid) |> round(2) |> 
  matrix(nrow = 3,byrow =T, dimnames = dmnms) |> 
  as.data.frame() |> 
  knitr::kable("html", digits=2, table.attr = "style='width:40%;'") |> 
  kableExtra::column_spec(column = 1, width = "5em") |> 
  kableExtra::column_spec(column = 2:11, width = "3em") |> 
  kableExtra::kable_styling(full_width = F) |> 
  kableExtra::footnote(footnote_as_chunk=T, general = "group 1 is never treated.")
```

However, if we use only the non-treated data to estimate the fixed effects, we get what we expect under the parallel trends assumption: after removing fixed effects, all that is left is treatment effects (easy to see with only one unit per group-time).

```{r}
#| echo: true
#| code-fold: true
#| code-summary: show residuals by group-time using only non-treated outcome data
#| label: show residuals by group-time using only non-treated outcome data
#| fig-cap: Residual outcome data by group-time, estimates only non-treated outomes

# Stage 1: Estimate fixed effects using ONLY UNTREATED observations
untreated_data4_rev <- dat4_rev[dat4_rev$treated == 0, ]
stage1 <- fixest::feols(outcome ~ 1 | group_id + period_id, data = untreated_data4_rev)
# compute residuals
dat4_rev$outcome_resid_gardner <- 
  dat4_rev$outcome - predict(stage1, newdata = dat4_rev)

dat4_rev |> dplyr::arrange(group_id) |> 
  dplyr::pull(outcome_resid_gardner) |> round(2) |> 
  matrix(nrow = 3,byrow =T, dimnames = dmnms) |> 
  as.data.frame() |> 
  knitr::kable("html",digits=2,table.attr = "style='width:40%;'") |> 
  kableExtra::column_spec(column = 1, width = "5em") |> 
  kableExtra::column_spec(column = 2:11, width = "3em") |> 
  kableExtra::kable_styling(full_width = F) |> 
  kableExtra::footnote(footnote_as_chunk=T, general = "group 1 is never treated.")
```

We can now regress these residuals against the treatment indicator and estimate the ATT:

```{r}
#| echo: true
#| code-fold: true
#| code-summary: estimating fixed effects with untreated only
#| label: estimating fixed effects with untreated only

# Stage 2: Regress residualized outcome on treatment
stage2 <- fixest::feols(outcome_resid_gardner ~ treated, data = dat4_rev)

cat("Gardner Two-Stage Estimate:", round(coef(stage2)[2], 6), "\n")
```

Which is the ATT as constructed in this example.

This two-stage DiD calculation is implemented in the `did2s` package, which we call as follows:

```{r}
#| echo: true
#| code-fold: true
#| code-summary: estimate treatment effects with 2sdid
#| label: estimate treatment effects with 2sdid
#| message: false
#| fig-width: 4
#| fig-align: "center"
#| output: asis
res <- did2s::did2s(
  data = dat4_rev
  , yname = 'outcome'
  , treatment = 'treated'
  , first_stage = ~ 0 | group_id + period_id
  , second_stage = ~i(treated, ref=FALSE)
  , cluster_var = 'group_id'
  , verbose = FALSE
)
#res
#wsx <- 
  fixest::etable(
  res, fitstat=c('n'), markdown = "images/" #tex = FALSE
) 

```

and did2s will also prepare an event study plot:

```{r}
#| echo: true
#| code-fold: true
#| code-summary: event study using 2sdid
#| label: event study using 2sdid
#| message: false

out <- did2s::event_study(
    data = dat4_rev |> 
      dplyr::mutate(
        g = dplyr::case_when(
          group_id == 1 ~ 0
          , group_id == 5 ~ 0
          , TRUE ~ 8
        )
      )
  , yname = 'outcome'
  , idname = 'group_id'
  , gname = 'g'
  , tname = 'period_id'
  , estimator = "did2s"
)
did2s::plot_event_study(out) + 
  theme_light(base_size = 18) + 
  theme(axis.title = element_text(size = 10, face = "bold"), legend.position="none")
```
