---
title: "The recipes package"
subtitle: "BSMM8740-2-R-2027F [WEEK - 2]"
author: "Dr. L.L. Odette"
footer: "[bsmm-8740-fall-2027.github.io/osb](https://bsmm-8740-fall-2027.github.io/osb/)"
logo: "images/logo.png"
format:
  revealjs:
    width: 1600
    height: 900
    chalkboard: true
    theme: slides_lec02.scss
    multiplex: true
    transition: fade
    slide-number: true
    margin: 0.05
    html-math-method: mathjax
editor: visual
execute:
  freeze: auto
  warning: false
  message: false
---

```{r setup}
#| include: false

library(knitr)
library(tidyverse)
library(recipes)
library(caret)
library(rsample)
library(skimr)
library(patchwork)

opts_chunk$set(comment = "", digits = 3, tidy = FALSE, prompt = TRUE,
               fig.align = "center")

# Palette — applied to all ggplot2 figures in this deck
pal <- list(
  green  = "#2D6A4F",
  sage   = "#52B788",
  terra  = "#E76F51",
  bg     = "#FAFAF8",
  text   = "#1A2332",
  code   = "#E8F4EC",
  gray   = "#6B7280",
  sand   = "#F4A261"
)

theme_kitchen <- function(base_size = 22) {
  theme_minimal(base_size = base_size) +
  theme(
    plot.background  = element_rect(fill = pal$bg,  color = NA),
    panel.background = element_rect(fill = pal$bg,  color = NA),
    panel.grid.major = element_line(color = "#E5E7EB"),
    panel.grid.minor = element_blank(),
    axis.text        = element_text(color = pal$text),
    axis.title       = element_text(color = pal$text, face = "bold"),
    plot.title       = element_text(color = pal$green, face = "bold"),
    plot.subtitle    = element_text(color = pal$gray),
    legend.position  = "top",
    legend.background = element_rect(fill = pal$bg, color = NA)
  )
}

# Helper: customer-product heatmap (used in slides 19 and 24)
hm <- function(M, title_txt) {
  tibble::as_tibble(M) |>
    dplyr::mutate(customer = row_number()) |>
    tidyr::pivot_longer(-customer, names_to = "product", values_to = "val") |>
    dplyr::mutate(product = as.integer(factor(product))) |>
    ggplot(aes(product, customer, fill = val)) +
    geom_tile() +
    scale_fill_gradient(low = pal$bg, high = pal$green) +
    scale_y_reverse() +
    coord_fixed() +
    labs(title = title_txt, x = "Product", y = "Customer", fill = "Purchases") +
    theme_kitchen(base_size = 16)
}

# Load data once for reuse across slides
Sacramento    <- modeldata::Sacramento
data(segmentationData, package = "caret")
seg_train <- segmentationData |> dplyr::filter(Case == "Train") |> dplyr::select(-Case)
seg_test  <- segmentationData |> dplyr::filter(Case == "Test")  |> dplyr::select(-Case)
dat <- datasets::airquality

set.seed(123)
M_biz <- matrix(rpois(300, lambda = 2), nrow = 20,
                dimnames = list(paste0("C", 1:20), paste0("P", 1:15)))
```

## Your test set is contaminated — and you don't know it yet {background-color="#FAFAF8"}

A junior analyst builds a model to predict customer churn.

She normalises all features with `scale()` on the full dataset — then splits into train and test.

The model scores 94% on test. It goes to production. Predictions are wrong.

**The bug:** `scale()` used test-set means and standard deviations to normalise the test set. The model had already seen the test distribution. This is *data leakage* — silent, common, and fatal in deployment.

::: muted
The formula method makes this easy to do accidentally. Recipes make it structurally impossible.
:::

## R formulas couple specification to execution — separating them requires gymnastics

```{r formula-demo}
#| echo: true
#| code-line-numbers: "1|2-4|5"
mod1 <- stats::lm(
  log(price) ~ type + sqft,
  data   = Sacramento,
  subset = beds > 2
)
```

One line does four things: subset rows, log-transform the outcome, dummy-encode `type`, and fit the model.

You cannot reuse "log-transform `price` and dummy-encode `type`" on new data without re-running the same formula — which means re-fitting on test-set statistics.

## Every package re-invents the formula because roles exceed what formulas support

| Package | What formula syntax had to add | Role that's missing |
|------------------------|------------------------|------------------------|
| `lme4` | `Reaction ~ Days + (Days \| Subject)` | **Random effect grouping** variable |
| `BradleyTerry2` | `~ reach[..] + (1\|\|..)` | **Competitor-specific** functions |
| `modeltools` | `diabetes ~ glucose \| pregnant + mass` | **Conditioning** variables |

R formulas know only two roles: *outcome* and *predictor*. Real models need at least six. Each package above re-invented the syntax from scratch.

##  {background-color="#2D6A4F"}

:::: {style="text-align: center; color: #FAFAF8; margin-top: 1.8em;"}
[A recipe is a plan, not an action]{style="font-size: 1.9em; font-weight: 700;"}

::: {style="font-size: 1em; color: rgba(255,255,255,0.75); margin-top: 0.6em;"}
Section A — Recipes separate planning from doing
:::
::::

## Recipes decouple *what you want* from *when you compute it*

:::::: workflow-row
::: {.w-box .w-green}
**recipe()**

[specify the plan]{.w-sub}
:::

[→]{.w-arrow}

::: {.w-box .w-sage}
**prep()**

[learn from training data]{.w-sub}
:::

[→]{.w-arrow}

::: {.w-box .w-terra}
**bake()**

[apply to any dataset]{.w-sub}
:::
::::::

. . .

::: {style="margin-top: 1.6em; font-size: 0.9em; color: #1A2332;"}
The three verbs stay the same whether you are applying the pipeline to the training set, test set, or new production data.
:::

## Three verbs replace the formula: `recipe()`, `prep()`, `bake()`

```{r recipe-basic}
#| echo: true
#| code-line-numbers: "1|3-5|8|10"
rec <- recipes::recipe(price ~ type + sqft, data = Sacramento)

rec <- rec |>
  recipes::step_log(price) |>
  recipes::step_dummy(type)

# learn parameters from training data
rec_trained <- recipes::prep(rec, training = Sacramento, retain = TRUE)
# apply to any dataset
design_mat  <- recipes::bake(rec_trained, new_data = Sacramento)
```

## `prep()` locks in training parameters; `bake()` applies them without leakage

The fix for the contaminated test set from slide 2:

```{r leakage-fix}
#| echo: true
#| eval: false
#| code-line-numbers: "1-2|4-5|7-8"
# WRONG — leaks test statistics into normalisation
scale(test_df$x, center = mean(test_df$x), scale = sd(test_df$x))

# RIGHT — uses only training statistics, applied to test
rec_prepped <- prep(rec, training = train_df)
bake(rec_prepped, new_data = test_df)

# The test set never touches prep() — leakage is structurally impossible
```

## dplyr-style selectors describe variables you haven't named yet

:::::: sel-grid
::: sel-card
**By type**

`all_numeric()`

`all_nominal()`

`all_integer()`
:::

::: sel-card
**By role**

`all_predictors()`

`all_outcomes()`

`has_role("strata")`
:::

::: sel-card
**By name pattern**

`contains(".R")`

`starts_with("PC")`

`ends_with("_ind")`
:::
::::::

. . .

::: {style="margin-top: 1.2em; font-size: 0.88em; color: #6B7280;"}
These resolve at `prep()` time — so they pick up dummy columns, PCA components, and discretized predictors whose names don't exist yet when you write the recipe.
:::

## Roles turn one recipe into a multi-purpose specification

```{r roles-demo}
#| echo: true
#| code-line-numbers: "1-2|3|4|5"
rec_roles <-
  recipes::recipe(Sacramento) |>
  recipes::update_role(price,      new_role = "outcome") |>
  recipes::update_role(type, sqft, new_role = "predictor") |>
  recipes::update_role(zip,        new_role = "strata")

summary(rec_roles)
```

##  {background-color="#2D6A4F"}

:::: {style="text-align: center; color: #FAFAF8; margin-top: 1.8em;"}
[Building a real pipeline]{style="font-size: 1.9em; font-weight: 700;"}

::: {style="font-size: 1em; color: rgba(255,255,255,0.75); margin-top: 0.6em;"}
Section C — airquality from raw data to model-ready matrix
:::
::::

## Ozone and Solar.R both have missings — but they need different treatments

```{r aq-skim}
#| echo: true
#| code-fold: true
#| code-summary: "skimr summary of airquality"
dat |> skimr::skim() |>
  dplyr::select(skim_variable, n_missing, numeric.mean, numeric.sd) |>
  gt::gt() |>
  gtExtras:::gt_theme_espn() |>
  gt::tab_options(table.font.size = gt::px(18)) |>
  gt::as_raw_html()
```

`Ozone`: 37 missing — impute from the mean of observed values. `Solar.R`: 7 missing — impute from its 3 nearest neighbours (K-NN).

## Step order in a recipe is the order of execution — sequence matters

```{r aq-split}
#| include: false
aq_split <- rsample::initial_split(dat, prop = 0.8)
aq_train <- rsample::training(aq_split)
aq_test  <- rsample::testing(aq_split)
```

```{r aq-recipe}
#| echo: true
#| code-line-numbers: "1-2|3-4|5-6|7-8|9-10"
aq_recipe <- recipes::recipe(Ozone ~ ., data = aq_train) |>
  recipes::update_role(Ozone, Solar.R, new_role = "NA_Variable") |>
  # impute Ozone by mean
  recipes::step_impute_mean(has_role("NA_Variable"), -Solar.R) |>
  # impute Solar.R by K-NN
  recipes::step_impute_knn(contains(".R"), neighbors = 3) |>
  # center then scale — AFTER imputation, so no NAs reach these steps
  recipes::step_center(all_numeric(), -has_role("NA_Variable")) |>
  recipes::step_scale(all_numeric(),  -has_role("NA_Variable"))
```

## `tidy()` makes every learned parameter visible and auditable

```{r aq-prep}
#| echo: true
aq_prep <- recipes::prep(aq_recipe, training = aq_train)
aq_prep |> recipes::tidy()
```

```{r aq-tidy3}
#| echo: true
aq_prep |> recipes::tidy(3)   # stored means from step_center
```

##  {background-color="#2D6A4F"}

:::: {style="text-align: center; color: #FAFAF8; margin-top: 1.8em;"}
[The step catalog]{style="font-size: 1.9em; font-weight: 700;"}

::: {style="font-size: 1em; color: rgba(255,255,255,0.75); margin-top: 0.6em;"}
Section D — six categories cover 90% of feature engineering
:::
::::

## Six categories cover 90% of feature engineering needs

| Category | What it does | Representative step |
|------------------------|------------------------|------------------------|
| **dummy** | Numeric encoding for qualitative predictors | `step_dummy()` |
| **zero-variance** | Remove constant or near-constant columns | `step_zv()`, `step_nzv()` |
| **impute** | Estimate missing values | `step_impute_knn()` |
| **decorrelate** | Reduce collinearity or dimensionality | `step_pca()` |
| **normalize** | Center and scale predictors | `step_normalize()` |
| **transform** | Symmetrize skewed predictors | `step_YeoJohnson()` |

::: muted
See the [recommended preprocessing table](https://www.tmwr.org/pre-proc-table) for which steps pair with which model families.
:::

## `step_date()` and `step_holiday()` turn one column into dozens of features

```{r date-recipe}
#| echo: true
#| eval: false
#| code-line-numbers: "1-3|4|5"
recipe(sales ~ date, data = sales_df) |>
  step_date(date, features = c("dow", "month", "week", "decimal")) |>
  step_holiday(date) |>   # US federal holiday indicators
  prep() |>
  bake(new_data = NULL)
```

One `Date` column becomes: `date_dow`, `date_month`, `date_week`, `date_decimal`, plus one binary column per holiday in the training window.

::: callout-note
This is where recipes earns the closing claim — **Recipes are the go-to for Date Engineering.** The formula method has no equivalent.
:::

##  {background-color="#2D6A4F"}

:::: {style="text-align: center; color: #FAFAF8; margin-top: 1.8em;"}
[Dimension reduction as a recipe step]{style="font-size: 1.9em; font-weight: 700;"}

::: {style="font-size: 1em; color: rgba(255,255,255,0.75); margin-top: 0.6em;"}
Section E — PCA composes with everything else
:::
::::

## High-dimensional data hides signal in noise — 300 cells, 15 products, one pattern

```{r biz-heatmap}
#| echo: false
#| fig-width: 8
#| fig-height: 5
#| fig-align: center
hm(M_biz, "Customer × Product purchases: raw data")
```

58 image features for cell segmentation. 15 product columns for a retailer. The structure is there — it's just buried. PCA finds it.

## PCA finds the direction of maximum spread in the data

```{r pca-geom}
#| echo: false
#| fig-width: 7
#| fig-height: 5
#| fig-align: center
set.seed(7)
x  <- rnorm(80)
y  <- 0.85 * x + 0.4 * rnorm(80)
df_pts <- tibble::tibble(x = x, y = y)
pc <- prcomp(df_pts)
d1 <- pc$rotation[, 1]
d2 <- pc$rotation[, 2]

ggplot(df_pts, aes(x, y)) +
  geom_point(color = pal$sage, alpha = 0.65, size = 2.5) +
  geom_segment(aes(x = 0, y = 0,
                   xend = d1[1] * 2.4, yend = d1[2] * 2.4),
               arrow = arrow(length = unit(0.3, "cm")),
               color = pal$green, linewidth = 1.6) +
  geom_segment(aes(x = 0, y = 0,
                   xend = d2[1] * 0.9, yend = d2[2] * 0.9),
               arrow = arrow(length = unit(0.3, "cm")),
               color = pal$terra, linewidth = 1.6) +
  annotate("text", x = d1[1] * 2.6, y = d1[2] * 2.4,
           label = "PC1\nmost variance", color = pal$green,
           fontface = "bold", size = 5, hjust = 0) +
  annotate("text", x = d2[1] * 1.0, y = d2[2] * 1.1 + 0.15,
           label = "PC2", color = pal$terra,
           fontface = "bold", size = 5) +
  coord_equal() +
  labs(title = "PCA rotates axes to align with variance",
       x = NULL, y = NULL) +
  theme_kitchen()
```

## `step_pca()` adds dimension reduction to any recipe in four lines

```{r seg-recipe}
#| echo: true
#| code-line-numbers: "1|2-4|5-7|9"
basic <- recipes::recipe(Class ~ ., data = seg_train) |>
  recipes::update_role(Cell, new_role = "ID") |>
  recipes::step_YeoJohnson(all_predictors()) |>
  recipes::step_normalize(all_predictors())

pca_rec <- basic |>
  recipes::step_pca(all_predictors(), num_comp = 5)

pca_prep <- recipes::prep(pca_rec, training = seg_train)
```

58 predictors → 5 principal components. PCA is just another step — it composes with everything before and after it.

## Five components out of 58 already separate poorly-segmented from well-segmented cells

```{r pca-scatter}
#| echo: false
#| fig-width: 7
#| fig-height: 5
#| fig-align: center
pca_baked <- recipes::bake(pca_prep, new_data = seg_test)

ggplot(pca_baked, aes(x = PC1, y = PC2, color = Class)) +
  geom_point(alpha = 0.45, size = 2) +
  scale_color_manual(values = c("PS" = pal$terra, "WS" = pal$green),
                     labels = c("PS" = "Poorly segmented",
                                "WS" = "Well segmented")) +
  labs(title = "PC1 vs PC2 separates cell classes",
       color = NULL) +
  theme_kitchen()
```

## The scree plot shows where additional components stop paying off

```{r scree}
#| echo: false
#| fig-width: 8
#| fig-height: 5
#| fig-align: center
set.seed(42)
n_cust <- 1000; n_prod <- 100; k_true <- 5
U_true <- matrix(rnorm(n_cust * k_true), n_cust, k_true)
V_true <- matrix(rnorm(n_prod * k_true),  n_prod, k_true)
svals  <- seq(5, 1, length.out = k_true)
Signal <- U_true %*% diag(svals, k_true) %*% t(V_true)
Noise  <- matrix(rnorm(n_cust * n_prod, sd = 0.8), n_cust, n_prod)
sv     <- svd(scale(Signal + Noise, center = TRUE, scale = FALSE))
cum_v  <- cumsum(sv$d^2) / sum(sv$d^2)

tibble::tibble(k = seq_along(cum_v), share = cum_v) |>
  dplyr::filter(k <= 20) |>
  ggplot(aes(k, share)) +
  geom_line(color = pal$gray) +
  geom_point(color = pal$green, size = 3) +
  geom_vline(xintercept = 5, linetype = "dashed", color = pal$terra) +
  annotate("text", x = 5.4, y = 0.25,
           label = "chosen k = 5\n(elbow)", color = pal$terra,
           size = 4.5, hjust = 0) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Cumulative variance explained by rank k",
       x = "Number of components (k)", y = "Variance explained") +
  theme_kitchen()
```

## SVD is the math underneath PCA: $M = U\Sigma V^\top$, keep the top $k$ singular values

```{r svd-heatmaps}
#| echo: false
#| fig-width: 10
#| fig-height: 4
#| fig-align: center
sv_biz <- svd(M_biz)
Mk <- function(k) {
  sv_biz$u[, 1:k, drop = FALSE] %*%
  diag(sv_biz$d[1:k], k) %*%
  t(sv_biz$v[, 1:k, drop = FALSE])
}
hm(Mk(2), "Rank 2 approximation") + hm(Mk(5), "Rank 5 approximation")
```

$U\Sigma$ = customer preferences in $k$ dimensions. $V$ = product features in $k$ dimensions. Rank 5 recovers the dominant purchase pattern from 300 numbers.

##  {background-color="#2D6A4F"}

:::: {style="text-align: center; color: #FAFAF8; margin-top: 1.8em;"}
[Extending and composing recipes]{style="font-size: 1.9em; font-weight: 700;"}

::: {style="font-size: 1em; color: rgba(255,255,255,0.75); margin-top: 0.6em;"}
Section F — building on what you've already computed
:::
::::

## Adding a step doesn't re-run earlier steps — computation is incremental

```{r extend}
#| echo: true
#| code-line-numbers: "1-3|5-6"
# rec_trained already has log(price) and step_dummy(type) prepped
standardized <- rec_trained |>
  recipes::step_center(recipes::all_numeric()) |>
  recipes::step_scale(recipes::all_numeric())

# only center and scale run — log() and dummy encoding are not recomputed
standardized <- recipes::prep(standardized)
summary(standardized)
```

This matters when earlier steps are expensive — e.g., K-NN imputation on wide data.

## A recipe wrapper separates the preprocessing pipeline from any model

```{r wrapper}
#| echo: true
#| code-line-numbers: "1|2|3-6"
lin_reg.recipe <- function(rec, data, ...) {
  trained <- recipes::prep(rec, training = data)
  lm.fit(
    x = recipes::bake(trained, new_data = data, all_predictors()),
    y = recipes::bake(trained, new_data = data, all_outcomes()),
    ...
  )
}
```

The model knows nothing about the preprocessing steps. Swap `lm.fit` for `glmnet::glmnet` and the recipe stays the same.

## Recipes solve the three formula limits

:::::: three-panel
::: p-card
**Extensible**

Nest any sequence of steps — including calls to Python, Weka, or TensorFlow. No re-implementation required.
:::

::: p-card
**Reusable**

`prep()` once on training data. `bake()` on the test set, cross-validation folds, or production — always using training statistics.
:::

::: p-card
**Efficient**

Add steps without recomputing expensive earlier steps. Incremental computation is built into the design.
:::
::::::

##  {background-color="#2D6A4F"}

::: {style="text-align: center; color: #FAFAF8; margin-top: 2.2em;"}
[Recipes are the go-to]{style="font-size: 2em; font-weight: 700; display: block;"}

[for Data Engineering.]{style="font-size: 2em; font-weight: 700; display: block; margin-top: 0.15em;"}
:::
